\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{enumitem}
\usepackage{geometry}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, decorations.pathreplacing, calc}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{url}
\geometry{margin=1in}

% doi command
\newcommand{\doi}[1]{\href{https://doi.org/#1}{\texttt{doi:#1}}}

% ------------------------------------------------------------------
% Theorem environments
% ------------------------------------------------------------------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ------------------------------------------------------------------
% Notation
% ------------------------------------------------------------------
\newcommand{\RC}{\mathcal{C}}          % configuration space
\newcommand{\RR}{\mathbb{Z}}           % integers
\newcommand{\id}{\mathrm{id}}          % identity configuration
\newcommand{\wt}[1]{w(#1)}            % weight

\title{A Process-Relational Foundation for Geometry}
\author{}
\date{}

\begin{document}

\maketitle

% ==================================================================
\begin{abstract}
Classical geometry takes space and points as primitives, treating metric
and positional properties as intrinsic to geometric objects. We argue
that these are not intrinsic but relational: they arise from explicit
comparisons with chosen reference structures. Accordingly, we develop a
foundational framework in which only finite relations and their
compositions are primitive. Finite composable sequences of relations,
called \emph{configurations}, represent processes. Objects arise as
equivalence classes under a \emph{congruence}---an equivalence relation
compatible with composition---that expresses observational
indistinguishability.

Our main results are:
\begin{itemize}
  \item \textbf{Theorem~\ref{thm:quotient} (Quotient Construction).}
        Any congruence on an associative relational system yields a
        well-defined quotient with associative composition.
  \item \textbf{Theorem~\ref{thm:metric} (Metric Emergence).}
        A weight function on a relational system with inverses
        induces a pseudometric---and under a separation condition, a
        genuine metric---on the quotient, derived from path
        optimization. When weights are integer-valued and $R$ is
        finite, the infimum is a minimum (Corollary~\ref{cor:discrete_metric}).
  \item \textbf{Theorem~\ref{thm:reconstruction} (Reconstruction).}
        Every finite metric space arises, up to isometry, from a
        relational system. The framework is universal for finite metric
        geometry.
  \item \textbf{Theorem~\ref{thm:equivalence} (Categorical Equivalence).}
        After identifying metrically equivalent relational systems, the
        metric realization functor yields an equivalence of categories
        between relational systems and finite metric spaces.
\end{itemize}
Concrete discrete constructions of the circle ($\RR/n\RR$), torus,
tree (negative curvature), and triangulated sphere (positive curvature)
illustrate the framework, along with a discrete relational account of
curvature via noncommutative composition. A dedicated section
establishes compatibility with classical geometry via Gromov--Hausdorff
convergence, framed as a bridge result rather than a foundational claim.
No continuity, limits, infinitesimals, or infinite processes are
assumed as foundational primitives; infinite discrete structures
such as the regular tree arise as derived objects from the framework's
finite rules, not as inputs.
\end{abstract}

% ==================================================================
\section{Introduction}
\label{sec:intro}

\subsection{The Classical Approach and Its Hidden Assumptions}

Standard geometry begins with a background space---typically Euclidean
$\mathbb{R}^n$---populated by points, and defines all further structure
(distance, angle, adjacency) on top of this substrate. This approach
has been enormously productive, but it embeds several assumptions that
are rarely made explicit.

First, \emph{position} is treated as an intrinsic property of a geometric
object, when in fact the position of any object is only meaningful
relative to a coordinate system or reference frame. A triangle does not
``have'' a location; it has a location \emph{relative to something else}.

Second, \emph{metric structure} is similarly treated as ambient and
pre-given, when it is in fact the result of a structural comparison
between two objects: the geometric object being measured and the
reference system supplying the unit. The length of a segment is not
a property of the segment alone, nor a property imposed by any
observer. It is what emerges at the interface between the segment
and its measurement context — produced by their contact, belonging
to neither one separately.

Third, the classical framework takes \emph{continuity and infinity} as
primitive, encoding them in the real number substrate. Yet no physical
measurement or finite computation ever directly accesses the continuum.
These are idealizations, and treating them as primitive conflates the
map with the territory.

\subsection{A Process-Relational Alternative}

We propose a foundational framework that takes none of these for granted.
The only primitives are \emph{finite relations} and their
\emph{compositions}. No background space, no coordinates, no points, no
continuous structures, and no infinite processes are assumed. All
geometric structure is derived from finite relational data.

The central move is this: \emph{objects are not primitive entities but
equivalence classes of finite relational processes}. A geometric object
is not a point or a set of points placed in space; it is the entire
class of processes that no relational test can distinguish. Space is not
assumed; it emerges as the structured set of such classes.

The framework rests on four explicit commitments:
\begin{enumerate}
  \item \textbf{Relations and compositions are primitive.} There is no
        substrate in which relations live; relations are the substrate.
  \item \textbf{Objects are equivalence classes of processes.} A
        geometric object is not a point or a set of points, but an
        equivalence class of finite composable sequences of relations
        under a congruence. This defines what an object \emph{is}.
  \item \textbf{Congruence encodes observational identity.} The
        criterion by which two processes are identified is that no
        relational test distinguishes them: composing any third
        configuration on either side produces the same result. This
        defines when two processes are the \emph{same} object.
  \item \textbf{All constructions are finite and discrete.} Continuity
        and infinity are not part of the framework; they are not
        approximated, they are simply absent.
\end{enumerate}

Commitment~4 is a foundational choice, not a limitation. Geometric
structure should be expressible in terms of finite relational data, or
not claimed to exist within this framework at all. The framework thereby
enforces a discipline: every geometric claim must be grounded in explicit,
finite, computable relational structure.

\subsection{Motivation}

Several independent developments across mathematics, physics, and
computation suggest that a discrete relational foundation for geometry
is not merely an intellectual exercise but a practical and physical
necessity.

In \emph{quantum gravity}, approaches including causal set
theory~\cite{causalset}, loop quantum gravity, and spin foam models
proceed from the premise that spacetime at the Planck scale is
fundamentally discrete and relational. The continuum is expected to be
an emergent approximation, not a primitive substrate. These frameworks
require a rigorous mathematical language for discrete relational
geometry that does not presuppose the smooth manifolds it is trying to
replace. The present framework offers one such language: whether it is
the right language for any specific quantum gravity model is an open
question, but it demonstrates that rigorous discrete relational geometry
is mathematically feasible without a continuous substrate.

In \emph{computational geometry and digital topology}, geometric
structures are always represented as finite discrete data. Classical
geometry must be discretized for computational use, introducing
approximation errors that must be managed. A framework that is
natively discrete requires no approximation: its objects are already
finite, its operations are already exact, and its equivalences are
already computable. The framework offers a foundational account of why
discrete computational geometry is not a degraded approximation of
the real thing but a fully legitimate geometric system in its own right.

In \emph{mathematical foundations}, the specific primitives of classical
geometry --- the point, the infinite line, the continuous circle --- are
not logically inevitable. They reflect the cognitive and historical
conditions under which geometry was first formalized: the experience of
space as pre-existing, of measurement as always available, of the
continuum as the natural habitat of number. A framework that derives
geometric objects from finite relational data, treating metric, position,
and scale as external contextual parameters rather than intrinsic
primitives, is not rejecting classical geometry. It is identifying which
of its features are necessary and which are contingent on the conditions
of its historical development.

In summary, the relational approach offers three principal advantages.
First, \emph{conceptual clarity}: by separating what is intrinsic to a
geometric object from what is supplied by the external context of
observation, the framework eliminates a class of conflations that have
persisted in classical foundations. Second, \emph{computational
tractability}: because all objects are finite and all operations are
exact, every geometric question reduces to a finite combinatorial
computation without approximation. Third, \emph{alignment with
discrete physical theories}: the framework provides a rigorous
mathematical language for spacetime models that treat discreteness and
relationality as primitive, rather than approximating them from a
continuous substrate.

\subsection{Main Contributions and Paper Structure}

The central results are the Quotient Construction
(\Cref{thm:quotient}), Metric Emergence (\Cref{thm:metric}),
Reconstruction of Finite Metric Spaces (\Cref{thm:reconstruction}),
and the Categorical Equivalence (\Cref{thm:equivalence}).

\Cref{sec:relational} defines relational systems and their directed
multigraph presentation. \Cref{sec:configurations} introduces
configurations as finite processes. \Cref{sec:congruence} defines
congruence. \Cref{sec:quotient} proves the quotient theorem and
establishes a complete classification of single-generator quotients,
including a partial classification of congruences.
\Cref{sec:metric} develops metric emergence in full generality,
with a corollary for the discrete case.
\Cref{sec:universality} establishes metric universality, minimal
presentations, failure modes, and the domain of validity.
\Cref{sec:examples} provides concrete constructions including a
full end-to-end worked example and negative, zero, and positive
curvature examples.
\Cref{sec:categorical} and \Cref{sec:higher} connect the framework to
category theory and higher structures. \Cref{sec:enriched} develops the
enriched categorical formulation and the equivalence theorem.
\Cref{sec:classical} establishes compatibility with classical geometry.
\Cref{sec:related} situates the work relative to existing frameworks,
including an explicit statement of what is new versus reformulated.
\Cref{sec:conclusion} summarizes and suggests directions.

\medskip
\noindent\textbf{Notation.} Throughout, $R$ denotes a set of primitive
relations, $\circ$ a partially defined composition, $\RC$ the set of all
configurations, and $\sim$ a congruence on $\RC$. We write $[C]$ for the
equivalence class of a configuration $C$, and $\RC/{\sim}$ for the
quotient. The empty configuration is denoted $\id$. The symbol $\sim$
is used consistently for a congruence; observational indistinguishability
is a \emph{property} relative to a given $\sim$, not a second relation
(see \Cref{def:obs-indist} and \Cref{rem:no-circularity}).

% ==================================================================
\section{Relational Systems}
\label{sec:relational}

\begin{definition}[Relational System]
\label{def:relsystem}
A \emph{relational system} is a pair $(R, \circ)$ where:
\begin{itemize}
  \item $R$ is a non-empty set of \emph{primitive relations},
  \item $\circ : R \times R \rightharpoonup R$ is a \emph{partially
        defined} composition.
\end{itemize}
\end{definition}

\begin{definition}[Compatibility]
\label{def:compat}
Relations $r_1, r_2 \in R$ are \emph{compatible} if $r_1 \circ r_2$
is defined.
\end{definition}

\begin{definition}[Associativity]
\label{def:assoc}
A relational system is \emph{associative} if, whenever all
compositions involved are defined,
\[
  (r_1 \circ r_2) \circ r_3 = r_1 \circ (r_2 \circ r_3).
\]
All relational systems in this paper are assumed associative.
\end{definition}

\begin{definition}[Relational Inverse]
\label{def:inverse}
A relation $r \in R$ has an \emph{inverse} $r^{-1} \in R$ if
\[
  r \circ r^{-1} = \id \quad\text{and}\quad r^{-1} \circ r = \id.
\]
A relational system is \emph{invertible} if every relation has an
inverse. In a system with congruence $\sim$, we say $r$ has an
\emph{inverse up to $\sim$} if $r \circ r^{-1} \sim \id$ and
$r^{-1} \circ r \sim \id$.
\end{definition}

\begin{remark}[Partial composition]
Partial composition is essential: requiring all pairs to compose would
force a uniformity inconsistent with most geometric and relational
structures of interest. The partial setting allows the framework to
encode directedness, type constraints, and locality naturally.
\end{remark}

\begin{remark}[Directed multigraph presentation]
\label{rem:multigraph}
A relational system $(R, \circ)$ admits an equivalent formulation as a
\emph{weighted directed multigraph} $G = (V, E, s, t, w)$, where $V$
is a set of vertices, $E$ a set of edges, $s, t : E \to V$ assign
source and target to each edge, and $w : E \to [0,\infty)$ is a
weight function. Primitive relations correspond to edges; compatibility
of $r_1$ and $r_2$ corresponds to $t(r_1) = s(r_2)$; composition
corresponds to concatenation of composable edges.

This formulation makes domain and codomain tracking explicit and
clarifies that the underlying structure is a \emph{free category on a
directed graph}, quotiented by the congruence $\sim$. In
\Cref{sec:universality} and \Cref{sec:enriched} we use this
presentation to state the universality and categorical equivalence
results precisely.
\end{remark}

% ==================================================================
\section{Configurations as Finite Processes}
\label{sec:configurations}

\begin{definition}[Configuration]
\label{def:configuration}
A \emph{configuration} is a finite composable sequence of relations:
\[
  C = (r_1, r_2, \dots, r_n), \quad r_i \in R,
\]
such that each consecutive pair $r_i, r_{i+1}$ is compatible
(Definition~\ref{def:compat}). The \emph{empty configuration} $\id$
(of length~$0$) is included as a two-sided identity.
The set of all configurations is denoted $\RC$.
\end{definition}

The term \emph{configuration} is used uniformly throughout. Configurations
represent finite, discrete processes or traversals in the relational
system. Informally, a configuration is a finite sequence of relational
steps that can be composed end-to-end.

\begin{definition}[Configuration Composition]
\label{def:confcomp}
For $C = (r_1, \dots, r_m)$ and $D = (s_1, \dots, s_n)$, their
\emph{composition} is:
\[
  C \circ D = (r_1, \dots, r_m, s_1, \dots, s_n),
\]
defined when $r_m$ and $s_1$ are compatible. We have
$\id \circ C = C \circ \id = C$ for all $C \in \RC$.
\end{definition}

\begin{remark}[Finiteness of configurations]
No limit, no infinite sequence, and no continuity condition is involved.
Every configuration is a finite object. Composition is concatenation,
a purely discrete, computable operation.
\end{remark}

% ==================================================================
\section{Congruence and Observational Indistinguishability}
\label{sec:congruence}

\begin{definition}[Congruence]
\label{def:congruence}
An equivalence relation $\sim$ on $\RC$ is a \emph{congruence} (with
respect to composition) if, for all configurations
$C_1, C_1', C_2, C_2' \in \RC$,
\[
  C_1 \sim C_1' \;\;\text{and}\;\; C_2 \sim C_2'
  \;\;\Rightarrow\;\;
  C_1 \circ C_2 \sim C_1' \circ C_2',
\]
whenever the compositions are defined.
\end{definition}

A congruence is the appropriate notion of equivalence for a system with
composition: it says that substituting equivalent configurations
preserves equivalence of the result. This is the only property of $\sim$
needed for Theorem~\ref{thm:quotient}.

\begin{definition}[Observational Indistinguishability]
\label{def:obs-indist}
Given a congruence $\sim$ on $\RC$, configurations $C_1$ and $C_2$ are
\emph{observationally indistinguishable (with respect to $\sim$)} if,
for every $D \in \RC$ compatible with both,
\[
  C_1 \circ D \sim C_2 \circ D
  \quad\text{and}\quad
  D \circ C_1 \sim D \circ C_2.
\]
\end{definition}

\begin{remark}[No circularity]
\label{rem:no-circularity}
Definition~\ref{def:obs-indist} does not define $\sim$; it describes a
property that a pair $(C_1, C_2)$ may or may not have \emph{relative to
a given} congruence $\sim$. The intended workflow is:
\begin{enumerate}
  \item Fix a relational system $(R, \circ)$.
  \item Specify $\sim$ by listing the relational laws of the system
        (e.g., $a^n \sim \id$).
  \item Verify that $\sim$ satisfies Definition~\ref{def:congruence}.
  \item Observe as a consequence that equivalent configurations are
        observationally indistinguishable.
\end{enumerate}
In practice, step~4 follows from steps~1--3 by the congruence property.
Observational indistinguishability is thus a \emph{derived} property,
not a primitive one.
\end{remark}

% ==================================================================
\section{Objects and the Quotient Construction}
\label{sec:quotient}

\begin{definition}[Object]
\label{def:object}
An \emph{object} is an equivalence class $[C]$ of a configuration $C$
under a congruence $\sim$:
\[
  [C] = \{ C' \in \RC \mid C' \sim C \}.
\]
\end{definition}

Objects are the geometric entities of the framework. A single geometric
``thing'' is not a configuration but the entire class of all configurations
that no \emph{relational test} can distinguish from one another.

\begin{theorem}[Quotient Construction]
\label{thm:quotient}
Let $(R, \circ)$ be an associative relational system with configuration
set $\RC$. Let $\sim$ be a congruence on $\RC$. Then:
\begin{enumerate}[label=(\roman*)]
  \item\label{qt:wdef} The quotient $\RC/{\sim}$ is well-defined as a
        set of equivalence classes.
  \item\label{qt:comp} Composition descends to a well-defined operation
        on $\RC/{\sim}$:
        \[
          [C_1] \circ [C_2] \;:=\; [C_1 \circ C_2].
        \]
  \item\label{qt:assoc} The induced composition is associative.
  \item\label{qt:id} The class $[\id]$ acts as a two-sided identity in
        $\RC/{\sim}$.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{\ref{qt:wdef}.} By definition, $\sim$ partitions $\RC$ into
disjoint, non-empty equivalence classes.

\textbf{\ref{qt:comp}.} We verify independence of representatives.
Suppose $[C_1] = [C_1']$ and $[C_2] = [C_2']$, i.e.,
$C_1 \sim C_1'$ and $C_2 \sim C_2'$. Since $\sim$ is a congruence
(Definition~\ref{def:congruence}),
\[
  C_1 \circ C_2 \sim C_1' \circ C_2',
\]
hence $[C_1 \circ C_2] = [C_1' \circ C_2']$. The operation is
independent of the choice of representatives.

\textbf{\ref{qt:assoc}.} For $C_1, C_2, C_3 \in \RC$, associativity
at the configuration level gives
$(C_1 \circ C_2) \circ C_3 = C_1 \circ (C_2 \circ C_3)$.
Applying part~\ref{qt:comp}:
\[
  ([C_1] \circ [C_2]) \circ [C_3]
  = [(C_1 \circ C_2) \circ C_3]
  = [C_1 \circ (C_2 \circ C_3)]
  = [C_1] \circ ([C_2] \circ [C_3]).
\]

\textbf{\ref{qt:id}.} For any $C \in \RC$,
$\id \circ C = C \circ \id = C$, hence $[\id] \circ [C] = [C] = [C]
\circ [\id]$.
\end{proof}

\begin{remark}[The quotient reveals, not creates]
The \emph{quotient} does not introduce structure. It stabilizes invariants
already present in $(R, \circ)$ by collapsing configurations that
are indistinguishable within the relational system.
Geometric structure is encoded in the generators $R$ and laws $\sim$;
the quotient merely makes that structure manifest.
\end{remark}

\subsection{Taxonomy of Single-Generator Quotients}
\label{sec:taxonomy}

The quotient construction produces different geometric objects
depending on the congruence. For single-generator systems, the
possibilities are completely classified. This classification is
foundational: it characterizes the expressive range of the simplest
relational systems and provides the algebraic criterion used
throughout the examples of \Cref{sec:examples}.

\begin{proposition}[Conditions for a well-formed closed object]
\label{prop:closed}
Let $(R, \circ)$ be an associative relational system with a single
generator $a$ and congruence $\sim$. The quotient $\RC/{\sim}$ is a
well-formed closed object if and only if the following two conditions
hold simultaneously:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Closure.} There exists $n \geq 1$ such that
        $a^n \sim \id$.
  \item \textbf{Uniqueness.} All intermediate steps are distinct:
        $a^j \not\sim a^k$ for all $0 \leq j < k < n$.
\end{enumerate}
When both hold, the quotient is $\mathbb{Z}/n\mathbb{Z}$.
When closure holds but uniqueness fails, the shape collapses to a
degenerate cycle. When closure fails entirely, the shape is open.
\end{proposition}

\begin{proof}
If $a^n \sim \id$ and all $a^0, \dots, a^{n-1}$ are pairwise
inequivalent, the $n$ classes $[a^0], \dots, [a^{n-1}]$ are distinct
and $\RC/{\sim} \cong \mathbb{Z}/n\mathbb{Z}$ by the congruence property.
If $a^j \sim a^k$ for some $j < k < n$, the quotient has period
$k-j < n$. If no $n$ satisfies $a^n \sim \id$, all powers are
distinct and the shape is open.
\end{proof}

\begin{lemma}[Reduction to minimal period]
\label{lem:reduction}
Let $(R, \circ)$ be an associative relational system with a single
generator $a$ and congruence $\sim$. If there exists $n \geq 1$ such
that $a^n \sim \id$ but the intermediate composites
$a^0, a^1, \dots, a^{n-1}$ are not all pairwise inequivalent, then
there exists a strictly smaller $k < n$ such that $a^k \sim \id$.
\end{lemma}

\begin{proof}
If intermediate composites are not all distinct, there exist indices
$i < j \leq n$ such that $a^i \sim a^j$. By the congruence property,
composing both sides on the right with $a^{-i}$ gives
\[
  \id \;\sim\; a^{i} \circ a^{-i} \;\sim\; a^j \circ a^{-i}
  \;=\; a^{j-i}.
\]
So $a^{j-i} \sim \id$ with $k := j - i < n$. Repeating this
reduction if necessary produces a minimal-period representative in
which all intermediate composites $a^0, a^1, \dots, a^{k-1}$ are
pairwise inequivalent.
\end{proof}

\begin{proposition}[Taxonomy of closure]
\label{prop:taxonomy}
Let $(R, \circ)$ be an associative relational system with a single
generator $a$ and congruence $\sim$. By Lemma~\ref{lem:reduction},
it suffices to classify systems at their minimal period, i.e., under
the assumption that all intermediate composites are distinct. Exactly
one of the following holds for such minimal-period systems:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Closed.} $a^n \sim \id$ for some $n \geq 1$ with
        all intermediate steps distinct. Quotient: $\mathbb{Z}/n\mathbb{Z}$.
  \item \textbf{Pseudo-periodic.} $a^n \sim \phi$ for some $n \geq 1$
        and $\phi \not\sim \id$, with all intermediate steps distinct.
  \item \textbf{Open.} No periodicity law holds. Quotient:
        $\mathbb{Z}$.
\end{enumerate}
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:reduction}, any system with repeated intermediate
composites reduces to one with strictly smaller period. It therefore
suffices to classify minimal-period systems.

For such systems, cases (i) and (ii) are distinguished by whether
the return target is $\id$ or some other $\phi \not\sim \id$. A
system cannot belong to both: if $a^n \sim \id$ then by definition
the return target is $\id$, not a nontrivial $\phi$.

Case (iii) covers all systems for which no finite composition returns
to any previously visited class; in this case all powers
$a^0, a^1, \dots$ are distinct. This is the negation of cases (i) and
(ii), so the three cases are mutually exclusive and exhaustive.
\end{proof}

\subsection{A Partial Classification of Congruences}
\label{sec:congruence_classification}

The paper's central claim is that geometry is determined by the
choice of congruence laws. It is therefore natural to ask: when do
two different congruences on the same relational system produce the
same geometric object? We give a complete answer for the
single-generator case.

We first define the key invariants intrinsically in relational
language, with no appeal to external symmetry notions.

\begin{definition}[Intrinsic period and shift order]
\label{def:shiftorder}
Let $R = \{a, a^{-1}\}$ and let $\sim$ be a congruence on $\RC$.

\begin{enumerate}[label=(\roman*)]
  \item The \emph{period} $n \in \mathbb{N} \cup \{\infty\}$ is the
        minimal $n > 0$ such that $a^n \sim \id$, if it exists, and
        $\infty$ otherwise.

  \item When $n < \infty$, define the \emph{loop subgroup}
        \[
          S \;:=\; \bigl\{ [m] \in \mathbb{Z}/n\mathbb{Z} \;\bigm|\;
          a^m \sim \id \bigr\} \;\subseteq\; \mathbb{Z}/n\mathbb{Z}.
        \]
        The \emph{shift order} is $k := n / |S|$.
\end{enumerate}
\end{definition}

\begin{lemma}
\label{lem:loopsubgroup}
The set $S$ in Definition~\ref{def:shiftorder} is a subgroup of
$\mathbb{Z}/n\mathbb{Z}$.
\end{lemma}

\begin{proof}
If $a^m \sim \id$ and $a^p \sim \id$, then by the congruence property
$a^{m+p} = a^m \circ a^p \sim \id \circ \id \sim \id$, so
$[m+p] \in S$. If $a^m \sim \id$, then composing both sides on the
right by $a^{-m}$ gives $a^m \circ a^{-m} \sim \id \circ a^{-m}$,
i.e., $\id \sim a^{-m}$, i.e., $a^{-m} \sim \id$, so $[-m] \in S$.
Hence $S$ is closed under addition and negation in
$\mathbb{Z}/n\mathbb{Z}$.
\end{proof}

\begin{remark}[Interpretation of $k$]
When $S = \{[0]\}$, no nontrivial power of $a$ is identified with
$\id$ within one period: the system is a pure cycle and $k = n$.
The case $S = \mathbb{Z}/n\mathbb{Z}$ — where every power of $a$
would be identified with $\id$ — is unreachable under the
minimal-period assumption of Proposition~\ref{prop:taxonomy}: by
Lemma~\ref{lem:reduction}, any such system reduces to period~1.
The relevant range is therefore $\{[0]\} \subseteq S \subsetneq
\mathbb{Z}/n\mathbb{Z}$, equivalently $1 < k \leq n$, with
$k = n$ corresponding to the closed (pure cycle) case and
$1 < k < n$ to the pseudo-periodic cases. Intermediate subgroups
yield intermediate values of $k$, encoding the internal loop
structure of the orbit. The shift order $k$ is the index
$[\mathbb{Z}/n\mathbb{Z} : S]$.
\end{remark}

\begin{proposition}[Complete congruence classification for
single-generator systems]
\label{prop:congclass}
Let $R = \{a, a^{-1}\}$ be the single-generator invertible system,
and let $\sim_1$, $\sim_2$ be congruences on $\RC$ with associated
invariants $(n_1, k_1)$ and $(n_2, k_2)$ as in
Definition~\ref{def:shiftorder}. The quotients $\RC/{\sim_1}$ and
$\RC/{\sim_2}$ are isomorphic if and only if one of the following
holds:

\begin{enumerate}[label=(\roman*)]
  \item $n_1 = n_2 = \infty$ (both open; both quotients are
        isomorphic to $\mathbb{Z}$);
  \item $n_1, n_2 < \infty$, $k_1 = k_2 = 1$, and $n_1 = n_2$
        (both closed with the same period);
  \item $n_1, n_2 < \infty$, $k_1, k_2 > 1$, and $n_1 = n_2$ and
        $k_1 = k_2$ (both pseudo-periodic with the same period and
        shift order);
  \item No other combination yields isomorphic quotients.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{Cases (i) and (ii).} Open congruences yield $\mathbb{Z}$;
closed congruences with period $n$ yield $\mathbb{Z}/n\mathbb{Z}$.
These are isomorphic as monoids if and only if the periods agree
($n_1 = n_2$). A finite monoid is never isomorphic to $\mathbb{Z}$.

\textbf{Case (iii).}
We prove both directions.

\textit{($\Rightarrow$) Isomorphic quotients imply equal $(n,k)$.}
If $\Phi : \RC/{\sim_1} \to \RC/{\sim_2}$ is an isomorphism, it
must send the generator $[a]_1$ to a generator of $\RC/{\sim_2}$.
Since $\RC/{\sim_2}$ is generated by $[a]_2$, the isomorphism
preserves the period: $a^{n_1} \sim_1 \id$ implies
$\Phi([a]_1)^{n_1} = \Phi([a^{n_1}]_1) = \Phi([\id]_1) = [\id]_2$,
so $a^{n_1} \sim_2 \id$, giving $n_2 \leq n_1$. By symmetry
$n_1 \leq n_2$, hence $n_1 = n_2 =: n$. Similarly, an isomorphism
maps the loop subgroup $S_1$ bijectively to $S_2$ (since it
preserves the identity and the group operation), so $|S_1| = |S_2|$
and thus $k_1 = n/|S_1| = n/|S_2| = k_2$.

\textit{($\Leftarrow$) Equal $(n,k)$ implies isomorphic quotients.}
Assume $n_1 = n_2 = n$ and $k_1 = k_2$. Since $k = n/|S|$ and
$S \leq \mathbb{Z}/n\mathbb{Z}$, equal $k$ with equal $n$ implies
$|S_1| = |S_2|$. Since $\mathbb{Z}/n\mathbb{Z}$ is cyclic, it
has exactly one subgroup of each order dividing $n$; therefore
$S_1 = S_2 =: S$.

Define the map $\Phi : \RC/{\sim_1} \to \RC/{\sim_2}$ by
$\Phi([a^j]_1) := [a^j]_2$.

\textit{Well-definedness.} Suppose $[a^j]_1 = [a^{j'}]_1$, i.e.,
$a^j \sim_1 a^{j'}$. By the congruence property,
$a^{j-j'} \sim_1 \id$, so $[j - j'] \in S_1 = S_2$, hence
$a^{j-j'} \sim_2 \id$, giving $a^j \sim_2 a^{j'}$ and
$[a^j]_2 = [a^{j'}]_2$.

\textit{Bijectivity.} The map $\Psi([a^j]_2) := [a^j]_1$ is the
inverse by the same argument with $\sim_1$ and $\sim_2$ swapped.

\textit{Composition-preservation.}
$\Phi([a^j]_1 \circ [a^m]_1) = \Phi([a^{j+m}]_1) = [a^{j+m}]_2
= [a^j]_2 \circ [a^m]_2 = \Phi([a^j]_1) \circ \Phi([a^m]_1)$.

Hence $\Phi$ is an isomorphism and $\RC/{\sim_1} \cong \RC/{\sim_2}$.

\textbf{Case (iv).} Across types, closed quotients (finite, no
nontrivial internal loops) are never isomorphic to pseudo-periodic
quotients (finite, with nontrivial $S$), nor to open quotients
(infinite).
\end{proof}

\begin{remark}[The invariant pair $(n, k)$ as a geometric signature]
The pair $(n, k)$ is a complete isomorphism invariant for
single-generator quotients: it determines the quotient up to
isomorphism and is determined by it. It encodes two independent
pieces of geometric information — $n$ measures the scale of the
orbit (how many steps before any return), and $k$ measures the
internal symmetry (how much loop structure is present within that
orbit). For multi-generator systems the analogous invariant is a
normal subgroupoid of the free path groupoid, which generalizes
both $n$ and $k$. A complete invariant theory for that case is the
central open problem of this framework.
\end{remark}

% ==================================================================
\section{Metric Emergence}
\label{sec:metric}

We show how a notion of distance arises from the relational system
without being imposed as a primitive. We state the general theorem
for real-valued weights first, then specialize to the integer-valued
case used throughout the finite discrete framework.

\begin{definition}[Weight Function]
\label{def:weight}
A \emph{weight function} is a map $w : R \to [0,\infty)$,
extended to configurations by:
\[
  w(C) = w(r_1) + \cdots + w(r_n)
  \quad \text{for } C = (r_1, \dots, r_n),
  \qquad w(\id) = 0.
\]
When $w$ takes values in $\mathbb{N}$, we call it a
\emph{discrete weight function}.
\end{definition}

\begin{definition}[Relational Path Between Objects]
\label{def:relpath}
A \emph{relational path from $[A]$ to $[B]$} is a configuration
$P \in \RC$ such that, for some representative $a \in [A]$,
the configuration $a \circ P$ belongs to $[B]$.
The set of all such paths is denoted $\mathrm{Path}([A],[B])$.
\end{definition}

\begin{theorem}[Metric Emergence]
\label{thm:metric}
Let $(R, \circ)$ be an invertible, associative relational system with
congruence $\sim$ and weight function $w : R \to [0,\infty)$.
Assume:
\begin{enumerate}[label=(A\arabic*)]
  \item\label{A1} $w(r) \geq 0$ for all $r \in R$,
  \item\label{A2} $w(C_1) = w(C_2)$ whenever $C_1 \sim C_2$
        (weight is a congruence invariant),
  \item\label{A3} $w(\id) = 0$ (the identity has zero cost),
  \item\label{A4} $w(r^{-1}) = w(r)$ for all $r \in R$
        (inverses have equal weight).
\end{enumerate}
Define:
\[
  d([A],[B]) = \inf\bigl\{ w(P) \;\bigm|\; P \in \mathrm{Path}([A],[B])
  \bigr\}.
\]
Then $d$ is a \emph{pseudometric} on $\RC/{\sim}$. If additionally:
\begin{enumerate}[label=(A\arabic*), start=5]
  \item\label{A5} $d([A],[B]) = 0 \Rightarrow [A] = [B]$
        (separation),
\end{enumerate}
then $d$ is a \emph{metric}.
\end{theorem}

\begin{proof}
\textbf{Non-negativity.} Immediate from~\ref{A1} and the definition
as an infimum over non-negative values.

\textbf{Self-distance.} $d([A],[A]) = 0$ because $\id \in
\mathrm{Path}([A],[A])$ and $w(\id) = 0$ by~\ref{A3}.

\textbf{Symmetry.} By invertibility and~\ref{A4}, reversing a path
$P = (r_1, \dots, r_n)$ from $[A]$ to $[B]$ yields
$P^{-1} = (r_n^{-1}, \dots, r_1^{-1})$, a path from $[B]$ to $[A]$
with $w(P^{-1}) = w(P)$. Hence $d([A],[B]) = d([B],[A])$.

\textbf{Triangle inequality.} Given $P \in \mathrm{Path}([A],[B])$
and $Q \in \mathrm{Path}([B],[C])$, the concatenation
$P \circ Q \in \mathrm{Path}([A],[C])$ satisfies
$w(P \circ Q) = w(P) + w(Q)$ by additivity of $w$. Taking infima:
\[
  d([A],[C]) \leq d([A],[B]) + d([B],[C]).
\]

\textbf{Metric.} Assumption~\ref{A5} upgrades the pseudometric to a
metric by providing separation.
\end{proof}

\begin{corollary}[Discrete metric emergence]
\label{cor:discrete_metric}
If additionally $w : R \to \mathbb{N}$ is a discrete weight function
and $R$ is finite with $w(r) \geq 1$ for all $r \in R$, then the
infimum in Theorem~\ref{thm:metric} is attained:
\[
  d([A],[B]) = \min\bigl\{ w(P) \;\bigm|\; P \in \mathrm{Path}([A],[B])
  \bigr\}.
\]
No limiting process or appeal to completeness is required.
\end{corollary}

\begin{proof}
When $w(r) \geq 1$ for all $r$, any path of $k$ steps has weight at
least $k$. For a fixed target weight $M$, the set of paths with
$w(P) \leq M$ is a finite set (bounded length, finite $R$). The
infimum over a finite non-empty set of non-negative integers is a
minimum.
\end{proof}

\begin{remark}[Scope of assumption \ref{A5}]
\label{rem:separation}
Assumption~\ref{A5} is not automatically satisfied. If $w$ assigns
weight zero to some non-identity relation $r$ with $r \not\sim \id$,
then $d([A],[A \circ r]) = 0$ even though $[A] \neq [A \circ r]$,
and $d$ is a pseudometric but not a metric. The separation condition
holds whenever every zero-weight configuration is equivalent to $\id$
under $\sim$. In particular, if $w(r) > 0$ for all $r \in R$, then
any non-empty path has positive weight and separation holds
automatically.
\end{remark}

\begin{remark}[Metric as ambient comparison, not observer-relative quantity]
\label{rem:ambient}
The framework's treatment of metric as an external parameter might
suggest that distances are observer-relative — that they depend on
who is measuring. This reading should be resisted. A distance in
this framework is not relative to a subject. It is the result of a
\emph{structural comparison} between two mathematical objects: the
relational system being measured and the reference system supplying
the weight function. Neither party in that comparison is a mind or a
perspective. Both are mathematical structures.

A useful analogy is thermal equilibrium in physics. Temperature is
not a property of a body alone, nor of the thermometer alone, nor of
a physicist performing the measurement. It is a property of the
contact between two physical systems at equilibrium. No subject is
required. The quantity emerges from the interface between the two
structures.

Length in this framework has the same character. A circle of
resolution $n = 100$ and one of resolution $n = 1000$ are the same
relational object. The different values of $n$ are not two observers
seeing the same thing differently — they are the results of two
different structural comparisons, each involving a different reference
step size. The length is neither intrinsic to the circle nor
externally imposed by a subject. It is produced at the interface
between the relational object and its measurement context.

This positions the framework at a point distinct from both classical
realism (length is intrinsic) and constructivism or observationalism
(length is what an observer measures). The correct description is
\emph{ambient comparison}: a quantity that belongs to the contact
between two structures, is fully determined by that contact, and
requires no subject to perform it. The resolution parameter is not
a cognitive choice — it is the scale at which the measurement context
operates. Changing it changes the comparison, not the object being
compared.
\end{remark}

% ==================================================================
\section{Metric Universality}
\label{sec:universality}

Theorem~\ref{thm:metric} shows that a relational system equipped with
a weight function produces a metric. We now establish the converse:
every finite metric space arises from a relational system. This
demonstrates that the framework is not merely generative but
\emph{universal} for finite metric geometry. The reconstruction
uses real-valued weights $w : R \to [0,\infty)$, the general case
of Definition~\ref{def:weight}.

\subsection{Reconstruction of Finite Metric Spaces}

\begin{theorem}[Reconstruction of finite metric spaces]
\label{thm:reconstruction}
Let $(X, d)$ be a finite metric space. Then there exists a relational
system $(R, \circ)$ with weight function $w : R \to [0,\infty)$ such
that the induced metric on $R/{\sim}$ is isometric to $(X, d)$.
\end{theorem}

\begin{proof}
We construct the relational system explicitly.

\textbf{Step 1: Construction.}
Set $R := \{ e_{xy} \mid x, y \in X \}$, with $e_{xy}$ representing
a direct relation from $x$ to $y$. Define composition by
$e_{xy} \circ e_{yz} = e_{xz}$ whenever the target of $e_{xy}$
matches the source of $e_{yz}$, i.e., when the middle vertex agrees.
Assign weights $w(e_{xy}) := d(x,y)$. The identity at each $x$ is
$e_{xx}$ with $w(e_{xx}) = d(x,x) = 0$.

\textbf{Step 2: Induced distance dominates $d$.}
The single-edge configuration $(e_{xy})$ connects $x$ to $y$ with
$w(e_{xy}) = d(x,y)$, so
\[
  d_{\mathrm{ind}}(x,y) \le d(x,y).
\]

\textbf{Step 3: $d$ dominates induced distance.}
For any configuration $\gamma = (e_{x_0 x_1}, \dots, e_{x_{n-1} x_n})$
from $x$ to $y$, repeated application of the triangle inequality gives
\[
  d(x,y) \le \sum_{i=0}^{n-1} d(x_i, x_{i+1}) = w(\gamma).
\]
Taking the infimum over all configurations, $d(x,y) \le d_{\mathrm{ind}}(x,y)$.

\textbf{Step 4: Equality and isometry.}
Combining Steps 2 and 3, $d_{\mathrm{ind}} = d$. Since $d$ is a
metric, $d(x,y) = 0 \Rightarrow x = y$, so the zero-distance
equivalence relation is trivial and $R/{\sim} \cong X$. The identity
map $X \to R/{\sim}$ is an isometry.
\end{proof}

\begin{remark}[Expressibility, not canonicity]
\label{rem:expressibility}
Theorem~\ref{thm:reconstruction} establishes \emph{expressibility}:
every finite metric space can be encoded as a relational system.
It does not establish \emph{canonicity}: there is no unique or
preferred encoding, and the choice of encoding is itself a contextual
decision.

This distinction is important for the paper's central philosophical
position. The introduction argues that metrics are external contextual
parameters, not intrinsic to geometric objects. The reconstruction
theorem is consistent with this position, but the consistency requires
care. In the proof, the metric $d(x,y)$ is assigned as an edge weight
$w(e_{xy})$ --- the metric appears as a primitive input to the
relational system. This does not mean the metric is intrinsic to the
geometric object; it means that a specific \emph{choice} of external
metric has been encoded into a specific relational presentation.
A different choice of metric would yield a different relational
presentation of the same underlying point set.

The correct interpretation is therefore: any metric can be expressed
relationally, but the relational expression always reflects a
contextual choice. The framework does not make the metric intrinsic
--- it makes explicit the choice that classical geometry leaves
implicit. Different relational presentations of the same point set,
with different weight functions, correspond to different observers
imposing different metrics from the external context.

This is also why Theorem~\ref{thm:minimal} matters: among all
relational presentations of a given metric, the minimal ones are
those in which no relation is redundant given the others. The minimal
presentation is the one that makes the contextual choice as lean
as possible --- retaining only the relations that the chosen metric
genuinely requires and cannot derive from composition of shorter
paths.
\end{remark}

\subsection{Minimal Presentations}

The construction in Theorem~\ref{thm:reconstruction} uses a complete
directed graph, which is highly redundant. Any metric space admits a
minimal relational presentation.

\begin{definition}[Generating relational system]
A relational system $(R, \circ, w)$ with underlying vertex set $X$
\emph{generates} a metric $d$ on $X$ if the induced metric
$d_{\mathrm{ind}}$ coincides with $d$.
\end{definition}

\begin{definition}[Redundant relation]
A relation $e_{xy} \in R$ is \emph{redundant} if there exists a
configuration $\gamma \in \mathrm{Conf}(x,y)$ not using $e_{xy}$
such that $w(\gamma) = d(x,y)$.
\end{definition}

\begin{theorem}[Minimal presentation]
\label{thm:minimal}
Every finite metric space $(X, d)$ admits a generating relational
system $(R, \circ, w)$ with no redundant relations, minimal with
respect to edge inclusion among all generating systems.
\end{theorem}

\begin{proof}
Start from the complete graph construction of
Theorem~\ref{thm:reconstruction}. Iteratively remove redundant
relations until none remain. Since $X$ is finite this process
terminates. Removing a redundant relation preserves the induced
metric (an alternative configuration of equal length exists). No
remaining relation can be removed without increasing some distance,
giving minimality.
\end{proof}

\begin{remark}[Failure modes and edge cases]
\label{rem:failuremodes}
We document the cases where the framework's machinery requires care.

\textit{Pseudometric rather than metric.} If the weight function
assigns zero weight to a non-identity relation $r \not\sim \id$,
the metric emergence theorem produces a pseudometric rather than a
metric (Remark~\ref{rem:separation}). This is not a failure — it is
the correct output for a non-separated relational system. The
separation condition (A5) is an explicit hypothesis, not a hidden
assumption.

\textit{Non-unique minimal presentations.} Minimal presentations are
not unique whenever the triangle inequality is slack at some triple.
If $d(x,z) < d(x,y) + d(y,z)$ for all intermediate points $y$,
then $e_{xz}$ is non-redundant and must appear in every minimal
presentation. If $d(x,z) = d(x,y) + d(y,z)$ for some $y$, then
$e_{xz}$ is redundant and may be omitted. Both choices are valid,
yielding different minimal presentations of the same metric. The
worked example of \Cref{sec:worked} illustrates the case where all
triangle inequalities are tight, producing a unique minimal path
presentation.

\textit{Pathological finite metric spaces.} Finite metric spaces
with no tight triangle inequalities (strictly ultra-metric spaces
where $d(x,z) \leq \max\{d(x,y), d(y,z)\}$ strictly) have a
unique minimal presentation consisting of all $|X|(|X|-1)$ directed
edges, since no multi-step path is ever as short as the direct edge.
These are the maximally complex cases: their minimal presentations
are complete directed graphs, and no compression is possible.
\end{remark}

\begin{remark}[Non-uniqueness of minimal presentations]
Minimal presentations are not unique in general. A simple example
illustrates this: the metric space on three points $\{a,b,c\}$ with
$d(a,b)=1$, $d(b,c)=1$, $d(a,c)=2$ (a metric line) has two minimal
presentations: one with edges $e_{ab}$ and $e_{bc}$ (the metric
line traversed left to right), and one where the roles of $b$-to-$a$
and $b$-to-$c$ are reversed. Both produce the same induced metric.

In general, the structure of minimal presentations reflects intrinsic
metric properties: the presence or absence of direct geodesic
relations, and the tightness of the triangle inequality at each
triple of points. Understanding the full relationship between
distinct minimal presentations of the same metric space --- in
particular, characterizing when two minimal presentations are
related by a natural transformation in $\mathbf{RelSys}_\sim$ ---
is an open problem. It is closely connected to the congruence
classification problem of the future directions.
\end{remark}

\subsection{Scope and Domain of Validity}
\label{sec:scope}

The theorems of this section establish the framework's reach for finite
metric spaces precisely. We state explicitly what is captured, what is
approximated, and what lies outside the framework.

\begin{theorem}[Domain of validity]
\label{thm:scope}
The process-relational framework captures the following classes of
metric spaces:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Finite metric spaces (exact).} By
        Theorem~\ref{thm:reconstruction} and the categorical
        equivalence of Theorem~\ref{thm:equivalence}, every finite
        metric space arises from a relational system, and the
        correspondence is an equivalence of categories. No
        approximation is involved.
  \item \textbf{Compact metric spaces (GH-approximate).} By
        Theorem~\ref{thm:ghbridge}, any compact metric space is the
        Gromov--Hausdorff limit of a sequence of finite relational
        systems. The classical space is recoverable as a limit but
        does not appear as a finite relational object.
  \item \textbf{Locally finite non-compact spaces (pointed
        GH-approximate).} By Theorem~\ref{thm:pointed_gh}, locally
        finite spaces such as the $k$-regular tree arise as pointed
        Gromov--Hausdorff limits of finite relational systems satisfying
        uniform local finiteness and rooted local convergence.
\end{enumerate}

The following lie outside the framework's current scope:
\begin{enumerate}[label=(\roman*)]
  \setcounter{enumi}{3}
  \item \textbf{Non-locally-finite infinite spaces.} Spaces without a
        uniform bound on the number of neighbours within any fixed
        radius do not satisfy the hypotheses of
        Theorem~\ref{thm:pointed_gh} and are not captured by the
        pointed construction.
  \item \textbf{Smooth Riemannian manifolds.} While compact Riemannian
        manifolds are GH-approximable by finite relational systems,
        their smooth structure — tangent spaces, differential forms,
        smooth maps — does not arise within the framework. Whether
        smooth structure can be encoded relationally is an open
        question.
  \item \textbf{Continuous symmetry groups.} Lie groups, $\mathrm{SO}(n)$,
        and the Poincar\'{e} group do not appear as objects of the
        theory; discrete cyclic approximations are available at any
        resolution but the continuous groups themselves require the
        real number substrate.
\end{enumerate}
\end{theorem}

\begin{proof}
Items (i)--(iii) are proved by the theorems cited. Items (iv)--(vi)
follow from the framework's foundational commitment to finite discrete
structure: they are classes that require either infinite vertex sets
without local finiteness bounds, or the real-number substrate, neither
of which is primitive in the framework.
\end{proof}

\begin{remark}[Boundary cases]
The boundary between cases (i)--(iii) and cases (iv)--(vi) is precise:
a metric space is exactly capturable if and only if it is finite; it is
GH-approximable if and only if it is compact or locally finite. The
remaining cases require structure that exceeds what finite relational
systems can provide. These are not vague incompleteness but identifiable
limits. In each case the limitation points to a concrete research
direction: extending the framework to length spaces, formalizing smooth
structure relationally, or characterizing which Lie group actions admit
discrete relational approximations.
\end{remark}

% ==================================================================
\section{Examples}
\label{sec:examples}

\subsection{The Discrete Circle}
\label{sec:circle}

Let $R = \{a\}$ with total composition $a \circ a^k = a^{k+1}$.
Configurations are:
\[
  \id,\; a,\; a^2,\; \dots,\; a^k, \;\dots
\]
Fix $n \geq 1$ and define the congruence generated by the single law:
\[
  a^n \sim \id.
\]
\textbf{Verification.} This is a congruence: if $a^k \sim a^{k'}$ and
$a^\ell \sim a^{\ell'}$, then $k \equiv k' \pmod{n}$ and
$\ell \equiv \ell' \pmod{n}$, so $k+\ell \equiv k'+\ell' \pmod{n}$,
giving $a^k \circ a^\ell \sim a^{k'} \circ a^{\ell'}$.

\textbf{Quotient.} By Theorem~\ref{thm:quotient}:
\[
  \RC/{\sim} = \{[a^0],[a^1],\dots,[a^{n-1}]\} \;\cong\;
  \mathbb{Z}/n\mathbb{Z},
\]
with $[a^k] \circ [a^\ell] = [a^{k+\ell \bmod n}]$.

\textbf{Geometric interpretation.} This cyclic structure is the
\emph{discrete circle of resolution $n$}. It encodes the defining
topological feature of the circle---closure under a uniform,
repeating process---purely through the relational law $a^n \sim \id$.
The $n$ objects $[a^0], \dots, [a^{n-1}]$ correspond to $n$ equally
distributed positions on the circle; the composition law encodes the
fact that stepping around the circle returns to the starting point.
No embedding in a plane, no arc length, and no angular coordinate
is used.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=1.2cm,
  every node/.style={draw, circle, inner sep=3pt, minimum size=7mm},
  arr/.style={-{Stealth[length=2mm]}, thick}
]
  \def\n{6}
  \def\radius{2cm}
  \foreach \i in {0,...,5} {
    \node (n\i) at ({90 - 360/\n * \i}:\radius)
      {\scriptsize $[a^{\i}]$};
  }
  \foreach \i [evaluate=\i as \j using {int(mod(\i+1,\n))}] in {0,...,5} {
    \draw[arr] (n\i) to[bend left=10] node[draw=none, midway,
      font=\scriptsize, outer sep=2pt] {$a$} (n\j);
  }
\end{tikzpicture}
\caption{The discrete circle $\mathbb{Z}/6\mathbb{Z}$ as a relational
quotient. Each node is an object (equivalence class); each arrow is
the generator~$a$. The relational law $a^6 \sim \id$ forces the
structure to close into a cycle. No spatial embedding is assumed.}
\label{fig:circle}
\end{figure}

\begin{remark}[Resolution is contextual, not intrinsic]
\label{rem:resolution}
The integer $n$ is not an intrinsic property of the circle. It
is a parameter expressing the \emph{resolution} of the relational
system relative to a chosen reference step. Two systems with different
values of $n$ represent the same relational shape observed at different
comparative granularities. The circle as a relational object is
the \emph{pattern}---closure under a uniform process---not a fixed
number of steps. For example, applying a reference step of one
centimetre to a circle of circumference one metre yields $n = 100$;
applying a reference step of one millimetre yields $n = 1000$. The
relational object --- the closure pattern --- is unchanged; only the
resolution of the external comparison differs.
\end{remark}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  every node/.style={draw, circle, inner sep=1pt, minimum size=5mm,
    font=\scriptsize},
  arr/.style={-{Stealth[length=1.5mm]}, thick}
]
  % n = 6
  \begin{scope}[xshift=0cm]
    \foreach \i in {0,...,5} {
      \node (na\i) at ({90 - 360/6 * \i}:1.3cm) {};
    }
    \foreach \i [evaluate=\i as \j using {int(mod(\i+1,6))}] in {0,...,5} {
      \draw[arr] (na\i) -- (na\j);
    }
    \node[draw=none, font=\small] at (0,-1.9cm) {$n = 6$};
  \end{scope}
  % n = 12
  \begin{scope}[xshift=4cm]
    \foreach \i in {0,...,11} {
      \node[inner sep=0.5pt, minimum size=3.5mm] (nb\i)
        at ({90 - 360/12 * \i}:1.3cm) {};
    }
    \foreach \i [evaluate=\i as \j using {int(mod(\i+1,12))}] in {0,...,11} {
      \draw[arr] (nb\i) -- (nb\j);
    }
    \node[draw=none, font=\small] at (0,-1.9cm) {$n = 12$};
  \end{scope}
  % n = 24
  \begin{scope}[xshift=8cm]
    \foreach \i in {0,...,23} {
      \node[inner sep=0pt, minimum size=2.5mm] (nc\i)
        at ({90 - 360/24 * \i}:1.3cm) {};
    }
    \foreach \i [evaluate=\i as \j using {int(mod(\i+1,24))}] in {0,...,23} {
      \draw[arr] (nc\i) -- (nc\j);
    }
    \node[draw=none, font=\small] at (0,-1.9cm) {$n = 24$};
  \end{scope}
\end{tikzpicture}
\caption{The same relational circle at three resolutions: $n = 6$,
$n = 12$, and $n = 24$. Each is the quotient $\mathbb{Z}/n\mathbb{Z}$
arising from the single law $a^n \sim \id$. The relational
identity---closure under a uniform process---is identical across all
three; only the resolution parameter $n$ differs. Increasing $n$
produces a finer discrete structure that more closely resembles a
smooth circle at any given scale of observation, but no limit is
taken: each is a complete, exact geometric object in its own right.}
\label{fig:resolution}
\end{figure}

\begin{remark}[Compositional pattern determines geometric identity]
\label{rem:identity}
The geometric identity of an object $[C] \in \RC/{\sim}$ is determined
entirely by the congruence $\sim$---that is, by the compositional pattern
of the relational laws. External parameters such as resolution $n$, scale,
and positional reference contribute nothing to this identity; they belong
to the external context of comparison, not to the object.

A direct consequence: in the discrete circle construction, the relational
law $a^n \sim \id$ and the cyclic structure $\mathbb{Z}/n\mathbb{Z}$ are
equivalent descriptions of the same geometric entity. Knowing the
compositional law determines the object; knowing the object up to
relational indistinguishability determines the law. The map from
congruence classes of relational laws to isomorphism classes of quotient
objects is a bijection.

This equivalence has a further consequence for geometric classification.
Two relational systems may be \emph{extensionally equivalent}---producing
isomorphic quotients---without being \emph{intensionally identical}---
having the same laws. Extensional equivalence is the geometric notion:
two systems are the same shape if no relational test distinguishes them.
Intensional identity is a property of the description. The gap between
the two is precisely where the classification of geometric structures
becomes non-trivial.
\end{remark}

\begin{example}[Distance on the discrete circle]
\label{ex:circle-distance}
We apply Theorem~\ref{thm:metric} to the discrete circle
$\mathbb{Z}/n\mathbb{Z}$. Assign unit weight to the generator:
$w(a) = 1$, so $w(a^k) = k$.

The distance between objects $[a^j]$ and $[a^k]$ is the minimum-weight
path between them. Two directed paths connect any two objects on the
cycle: the \emph{forward path} of length $|k - j| \bmod n$ steps, and
the \emph{backward path} of length $n - (|k - j| \bmod n)$ steps.
The metric selects the shorter:
\[
  d\bigl([a^j],[a^k]\bigr)
  \;=\; \min\!\bigl(\,|k-j| \bmod n,\;\; n - (|k-j| \bmod n)\,\bigr).
\]

For $n = 6$: the distance from $[a^0]$ to $[a^2]$ is $\min(2, 4) = 2$;
the distance from $[a^0]$ to $[a^5]$ is $\min(5, 1) = 1$, since the
backward step is shorter. The metric reflects the circular topology:
the maximum distance between any two objects is $\lfloor n/2 \rfloor$.

No coordinates, no arc length, and no embedding in a plane are used.
The metric is derived entirely from the relational weight function and
finite path optimization, as guaranteed by Theorem~\ref{thm:metric}.
\end{example}

\begin{corollary}[Circle as well-formed closed object]
\label{cor:circle-closed}
The discrete circle $\mathbb{Z}/n\mathbb{Z}$ is a well-formed closed
object in the sense of Proposition~\ref{prop:closed}: the congruence
$a^n \sim \id$ satisfies closure, and the $n$ classes
$[a^0], \dots, [a^{n-1}]$ are pairwise inequivalent (uniqueness).
Open shapes such as the line arise when no periodicity law is
imposed; see Proposition~\ref{prop:taxonomy}.
\end{corollary}

\subsection{End-to-End Worked Example: A Four-Point Metric Space}
\label{sec:worked}

We demonstrate the full pipeline of the framework on a concrete
finite metric space: presentation, reduction to minimal form, metric
reconstruction, and comparison with the classical description. This
example is chosen to be non-trivial — the minimal presentation is
not the complete graph — while remaining fully explicit.

\textbf{The metric space.} Let $X = \{p, q, r, s\}$ be four points
with distances:
\[
  d(p,q) = 1,\quad d(q,r) = 1,\quad d(r,s) = 1,\quad d(p,s) = 3,
\]
\[
  d(p,r) = 2,\quad d(q,s) = 2.
\]
This is the metric on four equally-spaced collinear points — a metric
line segment of length 3 with three unit steps.

\textbf{Step 1: Complete graph presentation.}
By Theorem~\ref{thm:reconstruction}, the canonical presentation uses
a directed complete graph on $\{p,q,r,s\}$ with twelve edges
$e_{xy}$ for every ordered pair $(x,y)$, weighted by $d(x,y)$.
This generates the correct metric but is maximally redundant.

\textbf{Step 2: Identifying redundant relations.}
A relation $e_{xy}$ is redundant if there exists an alternative
configuration of the same weight connecting $x$ to $y$.

Check $e_{pr}$: the path $e_{pq} \circ e_{qr}$ has weight
$1 + 1 = 2 = d(p,r)$. So $e_{pr}$ is redundant.

Check $e_{qs}$: the path $e_{qr} \circ e_{rs}$ has weight
$1 + 1 = 2 = d(q,s)$. So $e_{qs}$ is redundant.

Check $e_{ps}$: the path $e_{pq} \circ e_{qr} \circ e_{rs}$ has weight
$1 + 1 + 1 = 3 = d(p,s)$. So $e_{ps}$ is redundant.

By symmetry ($w(e_{xy}) = w(e_{yx})$), the reverse edges $e_{rp}$,
$e_{sq}$, $e_{sp}$ are redundant by the same argument. After removing
all six redundant edges, the minimal presentation contains exactly six
edges: $e_{pq}, e_{qp}, e_{qr}, e_{rq}, e_{rs}, e_{sr}$.

\textbf{Step 3: Verifying the minimal presentation.}
No edge in $\{e_{pq}, e_{qp}, e_{qr}, e_{rq}, e_{rs}, e_{sr}\}$ is
redundant. Each connects adjacent points with no shorter alternative
path between them (removing $e_{pq}$, for instance, would make $p$
unreachable from $q$ by a path of weight 1, since the only remaining
path from $p$ to $q$ uses $e_{qp}^{-1}$ — the same edge). The minimal
presentation is a path graph with unit-weight edges in both directions.

\textbf{Step 4: Metric reconstruction.}
The induced metric from the minimal presentation is:
\[
  d_{\mathrm{ind}}(x,y)
  = \min_{\gamma \in \mathrm{Conf}(x,y)} w(\gamma),
\]
computed by shortest-path over the six-edge graph. This gives the
same values as the original metric: $d_{\mathrm{ind}}(p,r) = 2$,
$d_{\mathrm{ind}}(p,s) = 3$, etc. Reconstruction is exact.

\textbf{Step 5: Classical comparison.}
Classically, four collinear points are described by specifying
coordinates — say $p = 0$, $q = 1$, $r = 2$, $s = 3$ on the real line
— and deriving distances as absolute differences. The relational
description is coordinate-free: no ambient line, no embedding, no
reference origin. The metric emerges entirely from the composition
structure of six primitive relations. The classical description
requires an ambient $\mathbb{R}$; the relational description requires
four objects and six rules.

\textbf{Observation on tightness.}
The triangle inequality is tight at every triple in this example:
$d(p,r) = d(p,q) + d(q,r)$, $d(q,s) = d(q,r) + d(r,s)$, and
$d(p,s) = d(p,q) + d(q,r) + d(r,s)$. Tightness at every triple
is exactly the condition under which all multi-step edges are
redundant — so tight triangle inequalities correspond precisely to
sparse minimal presentations. This illustrates the geometric meaning
of minimality: the presentation is minimal when the metric is fully
explained by the direct relations, with no shortcut distances needed.

\subsection{The Discrete Torus}
\label{sec:torus}

Let $R = \{a, b\}$. Impose the congruence generated by:
\[
  a \circ b \sim b \circ a,
  \quad a^n \sim \id,
  \quad b^m \sim \id.
\]

\textbf{Geometric interpretation.} The commutativity law
$a \circ b \sim b \circ a$ encodes two \emph{independent} relational
directions that do not interact. The periodicity laws $a^n \sim \id$
and $b^m \sim \id$ close each direction into a cycle. The combination
produces a two-dimensional discrete surface that wraps in both
directions simultaneously: a torus. No ambient $\mathbb{R}^3$, no
embedding, and no notion of ``surface'' in the classical sense is
used; the topology arises entirely from the relational laws.

By Theorem~\ref{thm:quotient}:
\[
  \RC/{\sim} \;\cong\; \mathbb{Z}/n\mathbb{Z} \;\times\;
  \mathbb{Z}/m\mathbb{Z}.
\]

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  arr/.style={-{Stealth[length=2mm]}, thick},
  node/.style={draw, circle, inner sep=2pt, minimum size=5mm,
    font=\scriptsize}
]
  \def\cols{4}
  \def\rows{3}
  % Draw nodes
  \foreach \i in {0,...,3} {
    \foreach \j in {0,...,2} {
      \node[node] (n\i\j) at (\i*1.6, \j*1.4)
        {\tiny $[a^\i b^\j]$};
    }
  }
  % Horizontal arrows (a direction), including wrap
  \foreach \i [evaluate=\i as \inext using {int(mod(\i+1,4))}]
    in {0,...,3} {
    \foreach \j in {0,...,2} {
      \ifnum\i<3
        \draw[arr, blue] (n\i\j) -- (n\inext\j);
      \else
        \draw[arr, blue, dashed] (n\i\j) to[bend left=40]
          node[draw=none, above, font=\tiny] {$a$} (n0\j);
      \fi
    }
  }
  % Vertical arrows (b direction), including wrap
  \foreach \j [evaluate=\j as \jnext using {int(mod(\j+1,3))}]
    in {0,...,2} {
    \foreach \i in {0,...,3} {
      \ifnum\j<2
        \draw[arr, red] (n\i\j) -- (n\i\jnext);
      \else
        \draw[arr, red, dashed] (n\i\j) to[bend left=40]
          node[draw=none, right, font=\tiny] {$b$} (n\i0);
      \fi
    }
  }
  % Legend
  \node[draw=none] at (2.4, -0.8) {
    \begin{tabular}{ll}
      \textcolor{blue}{---} & generator $a$ (horizontal) \\
      \textcolor{red}{---} & generator $b$ (vertical) \\
      \,{- - -} & wrap-around (periodic identification)
    \end{tabular}
  };
\end{tikzpicture}
\caption{The discrete torus $\mathbb{Z}/4\mathbb{Z} \times
\mathbb{Z}/3\mathbb{Z}$ as a relational quotient. Dashed arrows
indicate the periodic identifications $a^4 \sim \id$ (horizontal)
and $b^3 \sim \id$ (vertical). Commutativity $ab \sim ba$ means
the order of traversal is irrelevant. No embedding space is assumed.}
\label{fig:torus}
\end{figure}

\subsection{Relational Curvature via Noncommutativity}
\label{sec:curvature}

Let $R = \{a, b, a^{-1}, b^{-1}\}$ with
$a \circ a^{-1} \sim \id$ and $b \circ b^{-1} \sim \id$.
Suppose $a \circ b \not\sim b \circ a$, and define the
\emph{relational commutator}:
\[
  [a,b] \;=\; a \circ b \circ a^{-1} \circ b^{-1}.
\]

\textbf{Geometric interpretation.}
The commutator measures the discrepancy between two paths with the
same start and end: move first by $a$ then $b$, versus first by $b$
then $a$. If $[a,b] \sim \id$, the two paths are indistinguishable
and the relational structure is locally flat. If $[a,b] \not\sim \id$,
the two paths lead to observationally distinct outcomes: traversing
the loop $a \circ b \circ a^{-1} \circ b^{-1}$ does not return to the
starting state. This failure of closure is the discrete relational
analogue of \emph{holonomy}---the accumulated effect of parallel
transport around a closed loop---which in classical differential
geometry is the signature of nonzero curvature.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  arr/.style={-{Stealth[length=2.5mm]}, thick},
  node/.style={draw, circle, inner sep=3pt, minimum size=8mm}
]
  \node[node] (A) at (0,0)   {$P$};
  \node[node] (B) at (3,0)   {$Q$};
  \node[node] (C) at (3,2.5) {$R$};
  \node[node] (D) at (0,2.5) {$S$};
  \node[node] (E) at (1.5,1.25) [draw=none] {};

  % The square loop
  \draw[arr, blue]  (A) -- node[below]       {$a$}      (B);
  \draw[arr, red]   (B) -- node[right]       {$b$}      (C);
  \draw[arr, blue]  (C) -- node[above]       {$a^{-1}$} (D);
  \draw[arr, red]   (D) -- node[left]        {$b^{-1}$} (A);

  % Indicate failure to close
  \node[draw=none] at (1.5, 1.25) {$[a,b] \not\sim \id$};
  \draw[->, dashed, thick, gray] (A) to[out=30, in=150]
    node[above, draw=none, font=\small] {residual offset}
    ($(A)+(0.5,0.3)$);

  \node[draw=none, font=\small] at (1.5, -0.8) {
    Non-trivial commutator: the loop does not close.};
\end{tikzpicture}
\caption{Relational curvature as noncommutativity. The loop
$a \circ b \circ a^{-1} \circ b^{-1}$ travels from $P$ to $Q$ to $R$
to $S$ and back. If $[a,b] \not\sim \id$, the final state is
observationally distinct from the initial state: the loop has a
nontrivial holonomy, encoding discrete curvature. When $[a,b] \sim
\id$, the structure is locally flat (discrete torus case).}
\label{fig:curvature}
\end{figure}

Curvature in this framework is a discrete, finitely computable property:
check whether the commutator $[a,b]$ is equivalent to $\id$ under the
congruence. No differential calculus, no tangent spaces, and no
smoothness assumptions are required.

\begin{remark}[Relation to noncommutative geometry]
The use of noncommutativity to encode geometric curvature connects to
the program of noncommutative geometry initiated by Connes, where the
commutativity of coordinate algebras is relaxed to capture curved and
quantum spaces. In the present framework, the same idea appears in a
discrete, purely relational setting: the commutator $[a,b]$ is a
measurable relational observable, not an algebraic construction over
a function algebra. The framework thus makes the noncommutative
geometry intuition fully discrete and finite.
\end{remark}

\begin{remark}[Curvature as a special case of pseudo-periodicity]
\label{rem:curvature-pseudo}
The commutator $[a,b]$ is an instance of the pseudo-periodicity of
Proposition~\ref{prop:taxonomy}: traversing the loop
$a \circ b \circ a^{-1} \circ b^{-1}$ does not return to $\id$ but
to the shifted state $[a,b]$. This is case~(ii) of the taxonomy,
with $n = 4$ (the loop has four steps) and $\phi = [a,b]$.
Curvature is thus a two-generator instance of the same algebraic
structure that produces helices and spirals in the one-generator
case: a composition that shifts rather than closes.
\end{remark}

% ==================================================================
\subsection{Pseudo-Periodicity and the Taxonomy of Closure}
\label{sec:pseudoperiodic}

The examples so far — the circle, the torus, the open line — represent
two extremes of the closure spectrum: fully closed (periodicity law
present) and fully open (no periodicity law). A natural intermediate
category exists between them.

\begin{definition}[Pseudo-Periodic Relational System]
\label{def:pseudoperiodic}
A relational system $(R, \circ)$ with generator $a$ and congruence
$\sim$ is \emph{pseudo-periodic} with period $n$ and \emph{shift}
$\phi \in \RC/{\sim}$ if:
\[
  a^n \sim \phi, \quad \phi \not\sim \id.
\]
Each full traversal of $n$ steps does not return to the identity but
to a consistently shifted relational state $\phi$. The next traversal
begins from $\phi$, advances $n$ steps, and arrives at $\phi \circ
\phi = \phi^2$. Successive traversals accumulate the shift.
\end{definition}

The shift $\phi$ is a fixed relational transformation applied at the
end of each cycle. Its algebraic character determines the geometry of
the resulting object:

\begin{itemize}
  \item If $\phi$ encodes a \emph{translation} along an independent
        axis, the shape advances forward with each cycle. Combined with
        a closed oscillation, this produces a \emph{wave}: periodic
        in one direction, open in another.
  \item If $\phi$ encodes a \emph{scaling}, each cycle is larger than
        the previous by a constant factor. This produces a
        \emph{logarithmic spiral}: the same angular pattern repeated
        at increasing scale.
  \item If $\phi$ encodes a \emph{rotation} about a propagation axis,
        each cycle is rotated relative to the previous. This produces
        a \emph{helix}: constant angular advance per cycle along the
        rotation axis.
\end{itemize}

\begin{example}[Wave as product of closed and open systems]
\label{ex:wave}
Let $R = \{a, b, b^{-1}\}$ with $b \circ b^{-1} \sim \id$.
Impose the congruence generated by:
\[
  b^m \sim \id, \quad \text{no periodicity law on } a.
\]
The $b$-direction is closed with period $m$; the $a$-direction is open.
The quotient is:
\[
  \RC/{\sim} \;\cong\; \mathbb{Z} \;\times\; \mathbb{Z}/m\mathbb{Z}.
\]
This is an infinite sequence in the $a$-direction — the propagation
axis — combined with a finite cycle in the $b$-direction — the
oscillation axis. The result is a discrete wave: periodic in amplitude,
unbounded in propagation. The apparent infinity of the wave is entirely
in the $a$-direction and is, as with the line, the absence of a closure
law rather than the presence of an infinite object.
\end{example}

\begin{example}[Helix as pseudo-periodic system]
\label{ex:helix}
Let $R = \{a, b\}$. Impose:
\[
  a^n \sim b, \quad b^k \sim \id.
\]
The first law makes the system pseudo-periodic in $a$ with shift $b$.
The second closes the shift itself after $k$ accumulations. Each full
traversal of $n$ steps in $a$ advances one step in $b$; after $k$
such traversals, the $b$-direction closes. The quotient encodes a
discrete helix: $n$ steps per turn, $k$ turns before closure.
\end{example}

\begin{remark}[Pseudo-periodicity in the taxonomy]
Pseudo-periodic systems constitute the intermediate case of
Proposition~\ref{prop:taxonomy}: neither fully closed ($a^n \sim \id$)
nor fully open (no periodicity law), but shifting ($a^n \sim \phi$,
$\phi \not\sim \id$). The three cases of the taxonomy are mutually
exclusive and exhaustive for single-generator systems.
\end{remark}

\begin{remark}[Connection to semidirect products]
\label{rem:semidirect}
Pseudo-periodic relational systems correspond algebraically to
\emph{semidirect products}. If the shift $\phi$ acts on the cyclic
structure by an automorphism $\varphi$, the resulting quotient is the
semidirect product $\mathbb{Z}/n\mathbb{Z} \rtimes_\varphi \langle
\phi \rangle$. The helix, the screw motion, and the glide reflection
are all semidirect products of this form. The relational framework
recovers these classical symmetry structures as special cases of
pseudo-periodicity, without requiring the ambient space in which they
are classically defined.
\end{remark}

\begin{remark}[Continuous symmetries are absent by design]
\label{rem:nosymmetry}
The symmetry structures arising in this framework are discrete: cyclic
groups, finite semidirect products, and their combinations. Continuous
symmetry groups --- Lie groups, the rotation group $\mathrm{SO}(n)$,
the Poincar\'{e} group --- do not appear, because continuous structures
are absent from the framework by design.

This is a deliberate foundational choice, not an omission. Continuous
symmetries require the real number substrate and the associated
machinery of Lie theory. In the present framework, what classical
geometry describes as a continuous rotation is represented as a
discrete cyclic action at a specified resolution. The group
$\mathrm{SO}(2)$, for instance, is not an object in the framework;
the cyclic group $\mathbb{Z}/n\mathbb{Z}$ at resolution $n$ is. As $n$
increases, the discrete group approximates the continuous one more
finely, but the limit is never taken.

Whether this restriction is physically appropriate depends on the
domain of application. For quantum gravity and discrete spacetime
models, discrete symmetry groups are natural. For classical mechanics
and field theory, where continuous symmetries are central, the framework
provides a discrete approximation that may be refined to any required
precision but does not reproduce the continuum exactly. This is a known
and accepted limitation of any strictly finite framework.
\end{remark}

\begin{proposition}[Symmetry pattern from the generating function]
\label{prop:symmetry}
Let $(R, \circ)$ be a pseudo-periodic relational system with generator
$a$, period $n$, and shift $\phi$. The complete symmetry pattern of the
object $\RC/{\sim}$ is determined by $(n, \phi)$ alone, without
constructing or inspecting the object. Specifically:

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Orbit structure.} The shift $\phi$ acts on the $n$
        positions $[a^0], \dots, [a^{n-1}]$ of each cycle as a
        permutation $\sigma_\phi$. Object $[a^j]$ in cycle $c$ maps
        to $[\phi^c \circ a^j]$ in cycle $c+1$. The orbit of any
        position under successive cycles is directly computable from
        $\sigma_\phi$.

  \item \textbf{Higher-level periodicity.} If $\phi$ has finite order
        $k$ --- that is, $\phi^k \sim \id$ --- then the full object
        repeats after $k$ complete cycles, giving a higher-level
        period of $kn$ steps. If $\phi$ has infinite order, the object
        has no global periodicity but retains local symmetry within
        each cycle of $n$ steps.

  \item \textbf{Symmetry group.} The symmetry group of the object is:
        \[
          G \;\cong\; \mathbb{Z}/n\mathbb{Z} \;\rtimes_{\sigma_\phi}\;
          \langle \phi \rangle,
        \]
        where $\langle \phi \rangle \cong \mathbb{Z}/k\mathbb{Z}$ if
        $\phi$ has finite order $k$, and $\langle \phi \rangle \cong
        \mathbb{Z}$ otherwise. This group is fully determined by $n$,
        the order of $\phi$, and the permutation $\sigma_\phi$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{(i).} By definition of pseudo-periodicity, $a^n \sim \phi$.
The action of $\phi$ on $[a^j]$ is given by left composition:
$\phi \circ [a^j] = [\phi \circ a^j]$. Since $\phi$ is a fixed element
of $\RC/{\sim}$, this action is a well-defined endomorphism of the
$n$-cycle. That it is a permutation follows from the invertibility of
$\phi$ in an invertible relational system.

\textbf{(ii).} If $\phi^k \sim \id$, then after $k$ cycles the
accumulated shift returns to the identity: $[a^{kn}] \sim
[\phi^k \circ a^0] \sim [a^0]$. The object closes at step $kn$.
If no such $k$ exists, all cycles produce distinct relational states
and no global closure occurs.

\textbf{(iii).} The symmetry group is generated by $a$ (the
within-cycle step) and $\phi$ (the between-cycle shift), subject to
$a^n \sim \phi$ and the order of $\phi$. The semidirect product
structure follows from the action of $\phi$ on the $n$-cycle by
$\sigma_\phi$.
\end{proof}

\begin{corollary}[Compact encoding of pseudo-periodic objects]
\label{cor:encoding}
A pseudo-periodic relational object with period $n$, shift $\phi$ of
order $k$, and permutation action $\sigma_\phi$ is completely encoded
by the triple $(n, k, \sigma_\phi)$. The full object --- including its
symmetry group, orbit structure, and higher-level periodicity --- is
recoverable from this triple by the construction of
Proposition~\ref{prop:symmetry}.

In particular, the encoding cost is independent of the resolution
parameter: the same triple $(n, k, \sigma_\phi)$ encodes the object
at any resolution, with $n$ supplying only the within-cycle step count.
The intrinsic geometric information is contained in $k$ and
$\sigma_\phi$; the resolution $n$ is, as established in
Remark~\ref{rem:resolution}, an external contextual parameter.
\end{corollary}

\begin{remark}[Symmetry without construction]
\label{rem:symwithout}
Proposition~\ref{prop:symmetry} and Corollary~\ref{cor:encoding}
establish a property that is non-trivial from a classical perspective:
the complete symmetry pattern of a geometric object is readable
directly from its generating laws, without constructing the object,
without embedding it in a space, and without performing geometric
analysis. The symmetry is algebraic and the algebra is the object.

This is a direct consequence of the description/object equivalence
of Remark~\ref{rem:identity}: because the congruence and the object
are equivalent descriptions of the same thing, any property derivable
from the congruence is a geometric property of the object. Symmetry
is one such property. It is encoded in the laws before the object
exists.
\end{remark}

\begin{remark}[Connection to universality and categorical equivalence]
The compact encoding of Corollary~\ref{cor:encoding} situates
pseudo-periodic objects within the broader framework of
\Cref{sec:universality,sec:enriched}. The encoding triple
$(n, k, \sigma_\phi)$ determines a finite relational system whose
induced metric is a specific finite metric space; by
Theorem~\ref{thm:reconstruction}, this metric space is uniquely
recoverable from the relational data. Under the categorical
equivalence of Theorem~\ref{thm:equivalence}, the symmetry group
$G \cong \mathbb{Z}/n\mathbb{Z} \rtimes_{\sigma_\phi} \langle \phi
\rangle$ derived from the relational laws corresponds precisely to
the automorphism group of the induced metric space in
$\mathbf{LawMet}_{\mathrm{fin}}$.

Thus the pseudo-periodicity analysis is not an isolated result but
an instance of the general principle that intrinsic geometric
properties --- including symmetry --- are fully determined by the
relational presentation and can be read off without constructing
the geometric object explicitly.
\end{remark}

% ==================================================================
\subsection{Negative Curvature: The Infinite Regular Tree}
\label{sec:tree}

We construct a relational system whose induced metric exhibits
hyperbolic-type geometry: unique geodesics, exponential volume growth,
and 0-thin triangles. This is the canonical discrete model of negative
curvature.

Fix an integer $k \geq 2$. Define a relational system as follows.
The vertex set $V$ consists of all finite words over the alphabet
$\{1, \dots, k\}$, including the empty word $\varnothing$. For each
vertex $v$ and symbol $i \in \{1,\dots,k\}$, include a relation
$e_{v,i} : v \to vi$ (appending the symbol), and a reverse relation
$\bar{e}_{v,i} : vi \to v$ (removing the symbol). Assign unit weights
$w(e_{v,i}) = w(\bar{e}_{v,i}) = 1$.

This defines the infinite $k$-regular tree rooted at $\varnothing$.

\begin{proposition}
The induced metric satisfies:
\[
  d(v, w) = |v| + |w| - 2|v \wedge w|,
\]
where $|v|$ denotes word length and $v \wedge w$ is the longest
common prefix of $v$ and $w$.
\end{proposition}

\begin{proof}
Any configuration from $v$ to $w$ must ascend to the common ancestor
$v \wedge w$ and then descend. The shortest path does exactly this,
with length $(|v| - |v \wedge w|) + (|w| - |v \wedge w|)$. No shorter
path exists since the structure is a tree.
\end{proof}

\begin{theorem}[Hyperbolic features of the tree]
The relational system $(V, d)$ satisfies:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Unique geodesics.} Between any two vertices there is
        exactly one shortest configuration.
  \item \textbf{Exponential growth.} The number of vertices at
        distance $n$ from the root is $k^n$.
  \item \textbf{$0$-thin triangles.} For any three vertices, each
        side of the geodesic triangle is contained in the union of
        the other two.
\end{enumerate}
\end{theorem}

\begin{proof}
(i) The tree has no cycles, so there is a unique simple path between
any two vertices. (ii) Each vertex has exactly $k$ children, giving
$k^n$ vertices at depth $n$. (iii) In a tree, every geodesic triangle
is degenerate: one side lies on the path connecting the other two.
\end{proof}

\begin{remark}
Exponential volume growth and 0-thin triangles are the defining
combinatorial signatures of negative curvature in Gromov's sense.
This example shows that hyperbolic-type geometry arises from purely
relational data without any ambient curved space: the curvature is
a consequence of the branching structure of the relational
composition, not of any background geometry.
\end{remark}

% ==================================================================
\subsection{Positive Curvature: Triangulated Sphere}
\label{sec:sphere}

We construct a finite relational system exhibiting characteristic
features of positive curvature: reconverging geodesics and global
distance bounds. We work with a fixed finite triangulation rather
than a limit.

Let $T$ be any triangulation of the 2-sphere $S^2$ with vertex set
$V_T$ and edge set $E_T$. Define a relational system $(G_T, w_T)$:
vertices are elements of $V_T$; relations are edges of $E_T$ (taken
as bidirectional); weights are set to $w_T(e) = 1$ for all edges.

\begin{proposition}
In the relational system $(G_T, w_T)$:
\begin{enumerate}[label=(\roman*)]
  \item The induced metric $d_T$ counts the minimum number of
        triangulation edges in any path between two vertices.
  \item The diameter of $G_T$ is finite and bounded by the
        combinatorial structure of the triangulation.
\end{enumerate}
\end{proposition}

\begin{proof}
(i) Each edge has weight 1, so path length counts edges.
(iii) Any connected finite graph has finite diameter.
\end{proof}

\begin{example}[Geodesic non-uniqueness: the icosahedral graph]
\label{ex:icosahedron}
The icosahedral triangulation has 12 vertices, 30 edges, and 20
triangular faces. Each vertex is connected to 5 neighbours; the
icosahedral graph has diameter 3. By the icosahedral symmetry group
$A_5 \times \mathbb{Z}/2\mathbb{Z}$, every vertex has a unique
antipodal vertex at distance 3.

Let $v$ be any vertex and $\bar{v}$ its antipodal. There are exactly
6 distinct shortest configurations (paths of length 3) from $v$ to
$\bar{v}$, corresponding to the 6 pairs of opposite faces adjacent to
$v$. No shorter path exists since the diameter is 3.

This gives 6 distinct minimal configurations connecting the same pair
of objects — the relational analogue of multiple geodesics between
antipodal points on a positively curved surface.
\end{example}

\begin{remark}
The key distinction from the tree (negative curvature) is the
\emph{reconvergence} of paths: on the icosahedral graph, paths that
diverge from $v$ eventually reconverge at $\bar{v}$. Curvature in
the relational framework is the commutator structure of
Section~\ref{sec:curvature}; positive curvature manifests as the
global constraint that long ``straight'' paths cease to be minimal
beyond a certain scale, a property that emerges directly from the
relational composition rules of the triangulation.
\end{remark}

% ==================================================================
\subsection{Flat Geometry: Integer Grid}
\label{sec:grid}

We construct a finite relational system exhibiting flat (zero
curvature) geometry. Let $n \geq 1$. Define a relational system
$(G_n, w_n)$ as follows:
\begin{itemize}
  \item Vertex set $V_n := \{0, 1, \dots, n\}^2 \subset \mathbb{Z}^2$
        (a finite square grid);
  \item Edge set: connect each vertex $(i,j)$ to its eight
        neighbours --- four axis-aligned and four diagonal;
  \item Weight function: $w_n(e) := 1$ for all edges (axis and
        diagonal alike).
\end{itemize}

\begin{proposition}
The induced metric $d_n$ on $(G_n, w_n)$ is the Chebyshev
($\ell^\infty$) distance:
\[
  d_n\bigl((i_1,j_1),(i_2,j_2)\bigr)
  = \max\bigl\{|i_1 - i_2|,\; |j_1 - j_2|\bigr\}.
\]
\end{proposition}

\begin{proof}
Any path from $(i_1,j_1)$ to $(i_2,j_2)$ must cover a horizontal
displacement of $|i_1-i_2|$ and a vertical displacement of
$|j_1-j_2|$. Each step reduces at least one coordinate discrepancy
by 1. The minimum number of steps needed is therefore
$\max\{|i_1-i_2|, |j_1-j_2|\}$, achieved by using diagonal steps
to reduce both coordinates simultaneously. No shorter path exists.
\end{proof}

\begin{remark}[Flatness as commutative composition]
The Chebyshev metric on the integer grid is a flat discrete metric:
the two generators $a$ (horizontal) and $b$ (vertical) commute,
$a \circ b \sim b \circ a$, which is precisely the condition
$[a,b] \sim \id$ of \Cref{sec:curvature}. The torus of
\Cref{sec:torus} is this flat grid with periodic boundary
conditions, confirming that flatness is a property of the relational
composition rules, not of any ambient space.

The Chebyshev metric is one of several discrete flat metrics
constructible within the framework. Integer grids with axis-only
edges yield the $\ell^1$ (Manhattan) metric; the Chebyshev metric
arises from the full 8-connected grid. Both are exact, rational,
and require no approximation of irrational quantities.
\end{remark}

% ==================================================================
\section{Categorical Interpretation}
\label{sec:categorical}

The quotient construction has a natural categorical reading that
provides conceptual orientation before the formal development of
\Cref{sec:enriched}.

At an informal level:
\begin{itemize}
  \item Primitive relations in $R$ are \emph{generating morphisms}.
  \item The set of all configurations forms the \emph{free category}
        on $R$: objects are sources and targets of relations, and
        morphisms are composable chains.
  \item The congruence $\sim$ \emph{imposes relations} on this free
        category, quotienting it.
  \item The quotient $\RC/{\sim}$ is the \emph{category presented
        by generators $R$ and relations $\sim$} --- the categorical
        analogue of a monoid or group presentation.
\end{itemize}

When every generator has a compositional inverse up to $\sim$,
the quotient is a \emph{groupoid}. The circle yields
$\mathbb{Z}/n\mathbb{Z}$ as a one-object groupoid; the torus yields
$\mathbb{Z}/n\mathbb{Z} \times \mathbb{Z}/m\mathbb{Z}$ as a product.

The present framework \emph{contracts} rather than \emph{generalizes}
classical category theory: we start with relations alone and derive
objects, rather than assuming objects and morphisms as primitives.

The formal development --- including Lawvere metric spaces, the
metric realization functor, the universal property of the free
enriched category, and the equivalence of categories
$\mathbf{RelSys}_\sim \simeq \mathbf{LawMet}_{\mathrm{fin}}$ ---
is given in \Cref{sec:enriched}.

% ==================================================================
\section{Higher Relational Structure}
\label{sec:higher}

The framework extends naturally to higher levels.

\begin{itemize}
  \item \emph{Level~0.} Primitive relations in $R$ are $0$-cells.
  \item \emph{Level~1.} Configurations in $\RC$ are $1$-morphisms
        (composable chains of $0$-cells).
  \item \emph{Level~2.} Congruences between configurations
        are $2$-morphisms: identifications between processes.
  \item \emph{Level~$k$.} Congruences between $(k{-}1)$-morphisms
        are $k$-morphisms, for each $k \geq 2$.
\end{itemize}

When all morphisms at each level are invertible, this structure is a
\emph{strict $\omega$-groupoid}. When finiteness is maintained at each
level---that is, the generating data at every level is a finite set---
the entire hierarchy remains finite and discrete.

The discrete circle at Level~2 is the fundamental groupoid of a finite
directed graph, with configurations as $1$-morphisms and identifications
between configurations (homotopies in the discrete sense) as
$2$-morphisms.

\begin{example}[A 2-morphism on the discrete circle]
\label{ex:2morphism}
Consider the discrete circle $\mathbb{Z}/6\mathbb{Z}$ with generator
$a$. Two distinct configurations connect $[a^0]$ to $[a^4]$:
\[
  P \;=\; a^4 \quad (\text{four forward steps}),
  \qquad
  Q \;=\; a^{-2} \quad (\text{two backward steps}).
\]
Both $P$ and $Q$ are $1$-morphisms from $[a^0]$ to $[a^4]$ in the
groupoid. They are different processes but produce the same relational
outcome.

A \emph{$2$-morphism} $\alpha : P \Rightarrow Q$ is a relational
identification between these two processes: a witness that $P$ and $Q$
are indistinguishable as paths from $[a^0]$ to $[a^4]$. In the
discrete setting this identification is encoded by the congruence:
\[
  a^4 \;\sim\; a^{-2},
\]
which holds because $a^6 \sim \id$ implies $a^4 \circ a^2 \sim \id$,
so $a^4 \sim a^{-2}$.

\begin{center}
\begin{tikzpicture}[
  circ/.style={draw, circle, inner sep=2pt, minimum size=8mm,
    font=\scriptsize},
  arr/.style={-{Stealth[length=2mm]}, thick, gray, opacity=0.35},
  fwd/.style={-{Stealth[length=2.5mm]}, very thick, blue},
  bwd/.style={-{Stealth[length=2.5mm]}, very thick, red!70!black}
]
  \def\radius{2.2cm}
  \foreach \i in {0,...,5} {
    \node[circ] (n\i) at ({90 - 360/6 * \i}:\radius)
      {$[a^{\i}]$};
  }
  % Background arrows (gray)
  \foreach \i [evaluate=\i as \j using {int(mod(\i+1,6))}] in {0,...,5} {
    \draw[arr] (n\i) -- (n\j);
  }
  % Forward path P: [a^0]->[a^1]->[a^2]->[a^3]->[a^4]
  \draw[fwd] (n0) to[bend left=18] (n1);
  \draw[fwd] (n1) to[bend left=18] (n2);
  \draw[fwd] (n2) to[bend left=18] (n3);
  \draw[fwd] (n3) to[bend left=18] (n4);
  % Backward path Q: [a^0]->[a^5]->[a^4]
  \draw[bwd] (n0) to[bend right=18] (n5);
  \draw[bwd] (n5) to[bend right=18] (n4);
  % Labels
  \node[draw=none, blue, font=\small] at (-2.8cm, 0.5cm) {$P = a^4$};
  \node[draw=none, red!70!black, font=\small] at (2.8cm, 0.5cm)
    {$Q = a^{-2}$};
  % 2-morphism alpha label at centre
  \node[draw=none, font=\normalsize] at (0, 0)
    {$\alpha: P \Rightarrow Q$};
\end{tikzpicture}
\end{center}

The $2$-morphism $\alpha$ is the discrete relational analogue of a
homotopy between two paths: it certifies that the two processes are
equivalent without constructing a continuous deformation between them.
The forward path $P$ (blue, four steps clockwise) and the backward
path $Q$ (red, two steps anticlockwise) are algebraically identified
by the congruence $a^4 \sim a^{-2}$, which is the $2$-morphism. The
equivalence is algebraic, computable, and requires no ambient space.

\begin{remark}[Level structure of the 2-morphism]
In the higher categorical framework described above, a $2$-morphism
should properly be an element of a \emph{second-level} congruence
$\sim_2$ defined on the path category, distinct from the first-level
congruence $\sim$ that identifies configurations as objects. In the
present example, the first-level congruence $a^4 \sim a^{-2}$ serves
both roles simultaneously, because the cyclic group
$\mathbb{Z}/6\mathbb{Z}$ is simple enough that the two levels
collapse: every identification of configurations as objects is also an
identification of paths. A full treatment of the two-level structure
for more complex relational systems — where $\sim$ and $\sim_2$ are
genuinely distinct — is part of the higher categorical development
described in the future directions.
\end{remark}
\end{example}

% ==================================================================
\section{Enriched Categorical Formulation and Equivalence}
\label{sec:enriched}

The metric realization of a relational system has a natural
interpretation in enriched category theory: a relational system is
precisely a weighted directed graph, and the induced metric is the
hom-structure of the free category enriched over the monoidal poset
$([0,\infty], \geq, +, 0)$. This formulation provides the universal
characterization of the construction and yields a categorical
equivalence with finite metric spaces.

\subsection{Lawvere Metric Spaces}

\begin{definition}[Lawvere metric space~\cite{lawvere}]
A \emph{Lawvere metric space} is a set $X$ equipped with a function
$d : X \times X \to [0,\infty]$ satisfying, for all $x,y,z \in X$:
\[
  d(x,x) = 0,
  \qquad
  d(x,z) \le d(x,y) + d(y,z).
\]
Symmetry is not required in general. A classical metric space
corresponds to a symmetric Lawvere metric space with $d(x,y) = 0
\Rightarrow x = y$.
\end{definition}

\subsection{Relational Systems as Free Enriched Categories}

\begin{definition}[Free enriched category of a relational system]
Given a relational system $(R, \circ, w)$ presented as a weighted
directed graph $(G, w)$, define a category enriched over
$([0,\infty], \geq, +, 0)$ by:
\begin{itemize}
  \item Objects: vertices $V$ of $G$;
  \item Hom-value:
        $\mathcal{C}_G(x,y) := \min_{\gamma \in \mathrm{Conf}(x,y)} w(\gamma)$,
        where the minimum is taken over all configurations from $x$
        to $y$ (and equals $\infty$ if no such configuration exists).
\end{itemize}
Composition is given by addition: $\mathcal{C}_G(x,z) \leq
\mathcal{C}_G(x,y) + \mathcal{C}_G(y,z)$.
\end{definition}

\begin{proposition}
The assignment $(G,w) \mapsto \mathcal{C}_G$ defines a well-formed
$([0,\infty], \geq, +, 0)$-enriched category.
\end{proposition}

\begin{proof}
The triangle inequality for enriched hom-values holds because any
concatenation of configurations yields an admissible path:
$\mathcal{C}_G(x,z) \leq \mathcal{C}_G(x,y) + \mathcal{C}_G(y,z)$
by taking the concatenation of minimizing paths. The identity
condition $\mathcal{C}_G(x,x) = 0$ follows from the empty
configuration.
\end{proof}

\begin{theorem}[Free enriched category: universal property]
\label{thm:free_enriched}
The assignment $(G,w) \mapsto \mathcal{C}_G$ is left adjoint to the
forgetful functor from $([0,\infty], \geq, +, 0)$-enriched categories
to weighted directed graphs.
\end{theorem}

\begin{proof}[Sketch]
Given a weighted directed graph $(G,w)$ and an enriched category
$\mathcal{C}$, any weight-preserving graph map $\phi : G \to \mathcal{C}$
extends uniquely to an enriched functor $\mathcal{C}_G \to \mathcal{C}$
by sending each configuration to the composite of its image morphisms.
This is the universal property of the free construction.
\end{proof}

\begin{remark}
Theorem~\ref{thm:free_enriched} provides the universal characterization
requested by the framework's categorical foundations: the construction
of $\mathcal{C}_G$ from $(G,w)$ is not ad hoc but is the unique
functorial extension satisfying a universal mapping property. Geometry
arises as the hom-structure of a universally constructed enriched
category.
\end{remark}

\subsection{Metric Realization Functor}

\begin{definition}[Category of relational systems]
Let $\mathbf{RelSys}$ be the category whose objects are relational
systems $(G,w)$ and whose morphisms $F : (G,w) \to (G',w')$ are
graph morphisms preserving composition and satisfying
$w'(F(e)) \leq w(e)$ for all edges $e$.
\end{definition}

\begin{definition}[Category of Lawvere metric spaces]
Let $\mathbf{LawMet}$ denote the category of Lawvere metric spaces
with 1-Lipschitz maps.
\end{definition}

\begin{theorem}[Metric realization functor]
\label{thm:metric_functor}
There is a functor $\mathcal{M} : \mathbf{RelSys} \to \mathbf{LawMet}$
defined by $\mathcal{M}(G,w) := (V, d_G)$ where $d_G(x,y) =
\mathcal{C}_G(x,y)$.
\end{theorem}

\begin{proof}
For any morphism $F : (G,w) \to (G',w')$ and any configuration
$\gamma$ in $G$, the image $F(\gamma)$ is a configuration in $G'$
with $w'(F(\gamma)) \leq w(\gamma)$. Taking minima,
$d_{G'}(F(x), F(y)) \leq d_G(x,y)$, so $\mathcal{M}(F)$ is
1-Lipschitz. Functoriality with respect to composition and identities
is immediate.
\end{proof}

\subsection{Equivalence via Localization}

We now identify the precise sense in which relational systems and
metric spaces carry the same information.

\begin{definition}[Metric equivalence of relational systems]
Two relational systems $(G,w)$ and $(G',w')$ are \emph{metrically
equivalent} if there exists a bijection $\phi : V(G) \to V(G')$
with $d_{G'}(\phi(x), \phi(y)) = d_G(x,y)$ for all $x,y$.
Let $\mathbf{RelSys}_\sim$ denote the category obtained from
$\mathbf{RelSys}$ by identifying metrically equivalent objects.
\end{definition}

\begin{theorem}[Equivalence of categories]
\label{thm:equivalence}
The metric realization functor induces an equivalence of categories:
\[
  \mathbf{RelSys}_\sim \;\simeq\; \mathbf{LawMet}_{\mathrm{fin}},
\]
where $\mathbf{LawMet}_{\mathrm{fin}}$ is the full subcategory of
finite metric spaces.
\end{theorem}

\begin{proof}
\textbf{Essential surjectivity.} By Theorem~\ref{thm:reconstruction},
every finite metric space $(X,d)$ arises as $\mathcal{M}(G,w)$ for
an explicit relational system. Hence $\mathcal{M}$ is essentially
surjective onto $\mathbf{LawMet}_{\mathrm{fin}}$.

\textbf{Full faithfulness.}
We must show that for any two relational systems $(G,w)$ and $(G',w')$
with induced metrics $d_G$ and $d_{G'}$, the map
\[
  \mathbf{RelSys}_\sim\bigl((G,w),\,(G',w')\bigr)
  \;\longrightarrow\;
  \mathbf{LawMet}_{\mathrm{fin}}\bigl((V,d_G),\,(V',d_{G'})\bigr)
\]
is a bijection.

\textit{Injectivity.} If two morphisms $F_1, F_2 : (G,w) \to (G',w')$
induce the same 1-Lipschitz map on vertices, then $F_1(v) = F_2(v)$
for all $v \in V$, hence $F_1 = F_2$ as graph morphisms.

\textit{Surjectivity.} Let $f : (V, d_G) \to (V', d_{G'})$ be a
1-Lipschitz map. We construct a morphism of relational systems
realizing $f$.

By Theorem~\ref{thm:reconstruction}, both $(G,w)$ and $(G',w')$
are metrically equivalent (in $\mathbf{RelSys}_\sim$) to their
canonical complete-graph presentations: the relational system
$(\hat{G}, \hat{w})$ with vertex set $V$, a direct edge
$e_{xy} : x \to y$ for every ordered pair $(x,y) \in V \times V$,
and $\hat{w}(e_{xy}) = d_G(x,y)$; and similarly
$(\hat{G}', \hat{w}')$ with weights $\hat{w}'(e_{uv}') = d_{G'}(u,v)$.

Define a graph morphism $\hat{F} : \hat{G} \to \hat{G}'$ by
\[
  \hat{F}(v) = f(v), \qquad
  \hat{F}(e_{xy}) = e'_{f(x)f(y)}.
\]
This is well-defined since $\hat{G}'$ contains an edge between every
ordered pair. We verify the weight condition:
\[
  \hat{w}'\bigl(\hat{F}(e_{xy})\bigr)
  = \hat{w}'(e'_{f(x)f(y)})
  = d_{G'}(f(x), f(y))
  \leq d_G(x,y)
  = \hat{w}(e_{xy}),
\]
where the inequality holds because $f$ is 1-Lipschitz. Hence
$\hat{F}$ is a morphism in $\mathbf{RelSys}$.

Passing to $\mathbf{RelSys}_\sim$, we identify $(G,w)$ with
$(\hat{G},\hat{w})$ and $(G',w')$ with $(\hat{G}',\hat{w}')$, and
$\hat{F}$ is a morphism from the first to the second that induces
$f$ on metric spaces. This establishes surjectivity.

Therefore $\mathcal{M}$ is an equivalence.
\end{proof}

\begin{remark}[Relational systems as a presentation theory]
Theorem~\ref{thm:equivalence} is the categorical capstone of the
framework. It shows that relational systems and finite metric spaces
carry exactly the same information: a relational system is a
\emph{presentation} of a metric space, specifying generators and
composition rules from which the geometry is reconstructed. Different
relational systems presenting the same metric space are related by
metric equivalence — a form of gauge freedom in the relational
description. The geometry is the invariant content; the relational
system is one way of encoding it.
\end{remark}

% ==================================================================
\section{Connections to Existing Frameworks}
\label{sec:related}

The present framework shares surface similarities with several existing
theories, but differs from each in the foundational commitments of
\Cref{sec:intro}.

\paragraph{What is reformulated and what is genuinely new.}
A skeptical reader will ask whether the framework discovers new
phenomena or repackages known ones. Both occur, and the distinction
is worth making explicit.

\textit{Reformulated from known theory.} The quotient construction
(Theorem~\ref{thm:quotient}) is an instance of the universal
property of a category presented by generators and relations ---
this is standard categorical algebra~\cite{maclane}. The metric
emergence theorem (Theorem~\ref{thm:metric}) is, in the
single-object case, the construction of a word metric on a group
presentation; in the multi-object case it is the Cayley graph metric
of the quotient groupoid. Both are well-known objects. The categorical
equivalence (Theorem~\ref{thm:equivalence}) is a consequence of
Lawvere's identification of metric spaces with enriched
categories~\cite{lawvere} together with the reconstruction theorem.

\textit{Genuinely new.} The $(n,k)$ classification
(Proposition~\ref{prop:congclass}) --- a complete isomorphism
invariant for single-generator congruences defined purely in
relational language, with intrinsic period and shift order as the
invariants --- does not appear in the existing literature on
presentations of monoids or groups in this form. The pointed GH
emergence theorem (Theorem~\ref{thm:pointed_gh}) with its three
explicit conditions (uniform local finiteness, uniform edge bounds,
rooted local convergence) is a concrete relational formulation not
present in the metric geometry literature. The domain of validity
theorem (Theorem~\ref{thm:scope}), which draws the boundary between
exactly-capturable, GH-approximable, and out-of-scope spaces, is
new as an organized statement. And the philosophical framework ---
the explicit separation of intrinsic relational structure from
external contextual parameters (position, size, metric), the
expressibility/canonicity distinction, and the identification of
infinity as absence of closure rather than primitive object --- is an
original foundational position, not a restatement of existing work.

\textit{The honest summary.} The machinery is classical. The
organizational framework, the classification results, and the
philosophical position are new.

\paragraph{Category theory and enriched categories.}
The quotient is an instance of a category presented by generators and
relations~\cite{maclane, awodey}. The key difference is that classical
category theory takes objects as primitive data; here, objects are
derived. The framework is a \emph{generator-first} approach to
categorical structure. \Cref{sec:enriched} develops this further via
Lawvere's formulation of metric spaces as enriched
categories~\cite{lawvere}, yielding the universal property and
categorical equivalence results.

\paragraph{String rewriting systems.}
The relational system $(R, \sim)$ is formally a string rewriting
system, and the quotient is the monoid it presents~\cite{bookotto}.
The framework extends this by interpreting the presented structure
geometrically rather than computationally, and by allowing partial
rather than total composition.

\paragraph{Computational implementation.}
The framework is directly amenable to computational implementation.
Relational systems are finite graphs; configurations are directed paths;
congruences are equivalence relations on those paths, representable as
rewriting rules. Standard algorithms apply without modification:
shortest-path algorithms (Dijkstra, Bellman-Ford~\cite{clrs}) compute
the metric of Theorem~\ref{thm:metric} exactly; string rewriting normal
form algorithms (Knuth-Bendix completion~\cite{knuthbendix}) decide
congruence membership and classify quotient structures; graph isomorphism
algorithms~\cite{graphiso} determine extensional equivalence between
relational systems. A concrete computational advantage: geometric
equivalence checking --- determining whether two relational systems
describe the same shape --- reduces for single-generator systems to
comparing the pair $(n,k)$ by Proposition~\ref{prop:congclass}, a
computation requiring $O(n)$ time rather than the isometry testing
required in classical metric geometry. The finiteness and
discreteness of the framework are not restrictions on expressiveness ---
they are precisely the properties that make the geometry directly
computable. Every geometric question in the framework reduces to a
finite combinatorial computation with no approximation required.

\paragraph{Algebraic topology.}
The constructions of discrete circles, tori, and commutator-based
curvature connect to combinatorial algebraic topology~\cite{hatcher,
may}. The classical theory, however, builds on a continuous substrate
(simplicial complexes, topological spaces). The present framework
achieves analogous constructions without any continuous substrate.

\paragraph{Discrete and combinatorial geometry.}
Graph-based models of space~\cite{graphgeom} are structurally
compatible with the framework. The framework can be seen as a
compositional and algebraic enrichment of graph-theoretic geometry,
where edges (relations) carry algebraic structure and equivalence
classes (objects) replace vertices as the geometric units.

\paragraph{Directions toward physics.}
Several approaches to quantum gravity---causal set
theory~\cite{causalset}, loop quantum gravity, and spin foam
models---proceed from the premise that spacetime at the Planck scale
is discrete and relational rather than continuous and manifold-like.
The present framework is conceptually aligned with this premise and
provides a clean algebraic setting in which to formalize relational
spacetime structures.

As a concrete illustration, a causal set is a locally finite partially
ordered set $(X, \preceq)$ in which $x \preceq y$ means event $x$
causally precedes event $y$. In the present framework this maps
directly: the elements of $X$ are relational configurations, the
causal order $\preceq$ is the composition rule, and two causal
histories are equivalent if no causal test distinguishes them. The
quotient $\RC/{\sim}$ recovers the causal structure as a discrete
relational object, with curvature encoded in the commutator structure
of causal relations exactly as in \Cref{sec:curvature}. The framework
provides the algebraic machinery to reason about causal set geometry
--- symmetry, holonomy, equivalence --- without embedding the causal
set in a continuum.

Digital topology and computational geometry similarly operate with
finite, discrete data as primitive; the framework offers a
compositional foundation for these fields.

\paragraph{Relationship to classical geometry.}
The present framework does not approximate classical geometry as a
limit of finer relational structures. This distinction is important.
Taking a limit of discrete relational systems to recover, say, the
Euclidean plane would reintroduce the continuum through the back door
and would require the same limiting machinery the framework is designed
to avoid.

The correct relationship is different. Classical geometries --- Euclidean,
hyperbolic, spherical --- are not limiting cases of relational structures.
They are relational structures equipped with specific \emph{external
contextual parameters}: a choice of metric, a choice of reference frame,
a choice of scale. Remove those choices and the relational substrate
remains. Add them back and the classical geometry is recovered, not as
a limit but as a parameterization. Different choices of external
parameters over the same relational substrate produce different classical
geometries. The apparent multiplicity of classical geometries is a
consequence of the multiplicity of contextual parameterizations, not
a multiplicity of underlying structures.

A finer relational structure --- larger $n$ --- does produce a discrete
object that more closely resembles a classical smooth object at any
given scale of comparison. This is a practical observation about
approximation quality, not a mathematical limit: the discrete circle
$\mathbb{Z}/n\mathbb{Z}$ with $n = 10{,}000$ is observationally
indistinguishable from a smooth circle at most scales of physical
measurement, but it is not the limit of a sequence converging to the
smooth circle. The difference matters foundationally: approximation
quality is an external, contextual statement about resolution; the
smooth circle as a mathematical object requires actual infinity and
does not appear in the framework.

This means the framework is not weaker than classical geometry for
lacking limits; it is more general, in the sense that it separates
what is intrinsic to a geometric structure from what is supplied by
the context of observation.

% ==================================================================
\section{Compatibility with Classical Geometry}
\label{sec:classical}

The preceding sections establish the relational framework on its own
discrete terms. We now address the compatibility question: in what
precise sense do classical continuous geometric spaces relate to
relational systems? We show that classical spaces arise as
\emph{Gromov--Hausdorff limits} of sequences of finite relational
systems. This is a bridge result, not a foundational claim: the
classical spaces are not the targets the framework was aimed at, but
the framework is nonetheless compatible with them in a rigorous sense.

\subsection{Metric Approximation by Relational Systems}

\begin{definition}[Metric approximation]
Let $(X,d)$ be a compact metric space. A sequence of relational
systems $(G_n, w_n)$ with induced metrics $(V_n, d_n)$ and maps
$\phi_n : V_n \to X$ \emph{approximates} $(X,d)$ if:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Density.} $\phi_n(V_n)$ is $\varepsilon_n$-dense in $X$
        with $\varepsilon_n \to 0$;
  \item \textbf{Distance approximation.} $|d_n(x,y) - d(\phi_n(x),
        \phi_n(y))| \leq \varepsilon_n$ for all $x,y \in V_n$.
\end{enumerate}
\end{definition}

\begin{theorem}[Gromov--Hausdorff bridge theorem~\cite{buragoetal}]
\label{thm:ghbridge}
Let $(X,d)$ be a compact metric space. If $(G_n, w_n)$ approximates
$(X,d)$, then
\[
  d_{\mathrm{GH}}((V_n, d_n),\, (X,d)) \to 0,
\]
where $d_{\mathrm{GH}}$ denotes the Gromov--Hausdorff distance.
\end{theorem}

\begin{proof}
We construct an explicit correspondence between $(V_n, d_n)$ and
$(X, d)$ and bound its distortion.

\textbf{Forward map.} The given map $\phi_n : V_n \to X$ satisfies
$|d_n(x,y) - d(\phi_n(x), \phi_n(y))| \leq \varepsilon_n$ for all
$x, y \in V_n$.

\textbf{Reverse map.} Since $\phi_n(V_n)$ is $\varepsilon_n$-dense
in $X$, for each $z \in X$ there exists a point $\psi_n(z) \in V_n$
with $d(z, \phi_n(\psi_n(z))) \leq \varepsilon_n$. This defines a
map $\psi_n : X \to V_n$.

\textbf{Correspondence.} Define the correspondence
$\mathcal{R}_n \subseteq V_n \times X$ by
\[
  \mathcal{R}_n := \bigl\{ (x, \phi_n(x)) \mid x \in V_n \bigr\}
  \;\cup\;
  \bigl\{ (\psi_n(z), z) \mid z \in X \bigr\}.
\]
This is a correspondence: every $x \in V_n$ is related to
$\phi_n(x) \in X$, and every $z \in X$ is related to
$\psi_n(z) \in V_n$.

\textbf{Distortion bound.} For any pair $(x, z), (x', z') \in
\mathcal{R}_n$, we need to bound $d(\phi_n(x), z)$ and
$d(\phi_n(x'), z')$. There are two cases depending on which part
of the correspondence each pair comes from:
\begin{itemize}
  \item If $(x, z)$ comes from the first set, then $z = \phi_n(x)$,
        so $d(\phi_n(x), z) = 0 \leq \varepsilon_n$.
  \item If $(x, z)$ comes from the second set, then $x = \psi_n(z)$,
        so $d(\phi_n(x), z) = d(\phi_n(\psi_n(z)), z) \leq \varepsilon_n$
        by the density condition on $\psi_n$.
\end{itemize}
In both cases $d(\phi_n(x), z) \leq \varepsilon_n$, and likewise
$d(\phi_n(x'), z') \leq \varepsilon_n$. The triangle inequality then gives:
\begin{align*}
  \bigl| d_n(x,x') - d(z,z') \bigr|
  &\leq \bigl| d_n(x,x') - d(\phi_n(x),\phi_n(x')) \bigr| \\
  &\quad + \bigl| d(\phi_n(x),\phi_n(x')) - d(z,z') \bigr| \\
  &\leq \varepsilon_n + d(\phi_n(x), z) + d(\phi_n(x'), z') \\
  &\leq \varepsilon_n + \varepsilon_n + \varepsilon_n
  = 3\varepsilon_n.
\end{align*}

\textbf{Conclusion.} The Gromov--Hausdorff distance satisfies
$d_{\mathrm{GH}}((V_n,d_n),(X,d)) \leq \frac{3}{2}\varepsilon_n$
(by the characterization in terms of correspondences, see
\cite{buragoetal} Theorem~7.3.25). Since $\varepsilon_n \to 0$,
we obtain $d_{\mathrm{GH}}((V_n,d_n),(X,d)) \to 0$.
\end{proof}

\begin{corollary}
The following classical spaces arise as Gromov--Hausdorff limits of
sequences of finite relational systems via Theorem~\ref{thm:ghbridge}:
\begin{enumerate}[label=(\roman*)]
  \item The circle $(S^1, d_{S^1})$, via the cyclic graphs
        $(\mathbb{Z}/n\mathbb{Z}, d_n)$ with $d_n$ the arc-step metric;
  \item The sphere $(S^2, d_{S^2})$, via refining geodesic
        triangulations;
  \item The Euclidean plane $(\mathbb{R}^2, \|\cdot\|_2)$, locally, via
        integer grids with diagonal edges.
\end{enumerate}
Tree-like (hyperbolic) spaces are non-compact and require the
pointed extension; see Theorem~\ref{thm:pointed_gh} and
Example~\ref{ex:tree_pointed_gh} below.
\end{corollary}

\subsection{Pointed Gromov--Hausdorff Convergence}

The compactness assumption in Theorem~\ref{thm:ghbridge} excludes
non-compact limits such as the $k$-regular tree. The pointed
Gromov--Hausdorff framework extends compatibility to this case by
working with finite balls around a basepoint rather than the full
space.

\begin{definition}[Rooted relational system]
A \emph{rooted relational system} is a triple $(G, w, v)$ where
$(G, w)$ is a relational system and $v \in V(G)$ is a distinguished
basepoint.
\end{definition}

\begin{definition}[Pointed Gromov--Hausdorff convergence]
A sequence of rooted relational systems $(G_n, w_n, v_n)$ with
induced metrics $(V_n, d_n, v_n)$ \emph{converges in the pointed
Gromov--Hausdorff sense} to a pointed metric space $(V, d, v)$ if,
for every $R > 0$, the metric balls $B_{d_n}(v_n, R)$ converge in
the Gromov--Hausdorff sense to $B_d(v, R)$.
\end{definition}

\begin{theorem}[Pointed GH emergence]
\label{thm:pointed_gh}
Let $(G_n, w_n, v_n)$ be a sequence of rooted relational systems.
Assume:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Uniform local finiteness.} There exists $D \in \mathbb{N}$
        such that every vertex in every $G_n$ has degree at most $D$;
  \item \textbf{Uniform edge bounds.} There exist $0 < a \leq b < \infty$
        such that $a \leq w_n(e) \leq b$ for all $n$ and all edges $e$;
  \item \textbf{Rooted local convergence.} For every $R > 0$, the
        balls $B_{d_n}(v_n, R)$ stabilize up to isomorphism as
        weighted graphs for all sufficiently large $n$.
\end{enumerate}
Then there exists a pointed metric space $(V, d, v)$ such that
$(V_n, d_n, v_n)$ converges to $(V, d, v)$ in the pointed
Gromov--Hausdorff sense.
\end{theorem}

\begin{proof}[Sketch]
By uniform local finiteness and uniform edge bounds, the metric balls
$B_{d_n}(v_n, R)$ are uniformly controlled in size and geometry for
each fixed $R$. By rooted local convergence, they stabilize as
weighted graphs for large $n$. The stabilized balls are consistent
across radii: since $B_{d_n}(v_n, R)$ is the restriction of
$B_{d_n}(v_n, R+1)$ to vertices at distance at most $R$ from $v_n$,
once both balls have stabilized the restriction of the radius-$(R+1)$
limit to radius $R$ is identical to the radius-$R$ limit, so the
limiting family nests compatibly. Passing to the limit over
$R = 1, 2, 3, \dots$ via a diagonal argument then yields a
well-defined consistent family of finite metric spaces that assembles
into a pointed metric space $(V, d, v)$. Convergence on every finite
radius holds by construction.
\end{proof}

\begin{example}[The $k$-regular tree as a pointed GH limit]
\label{ex:tree_pointed_gh}
Let $T_k$ be the $k$-regular tree with root $v_0$ and unit edge
weights. Define $G_n$ to be the finite induced subgraph of $T_k$ on
the ball of radius $n$ about $v_0$, with $v_n := v_0$ and weights
inherited from $T_k$. Each $G_n$ is a finite rooted relational system.

The sequence $(G_n, w_n, v_n)$ satisfies all three hypotheses of
Theorem~\ref{thm:pointed_gh}: the tree is $k$-regular (condition~(i)),
all edges have weight~1 (condition~(ii)), and for any fixed $R$, the
ball $B_{d_n}(v_n, R)$ stabilizes to $B_{T_k}(v_0, R)$ for all
$n \geq R$ (condition~(iii)). The limit is therefore $(T_k, d, v_0)$,
the full $k$-regular tree with its canonical metric. Tree-like
(hyperbolic) geometry thus arises as a pointed Gromov--Hausdorff limit
of finite relational systems, even though it is not compact.
\end{example}

\subsection{The Compatibility Principle}

\begin{remark}[Parameterization, not approximation]
Theorem~\ref{thm:ghbridge} demonstrates that classical spaces are
compatible with the relational framework --- they are recoverable from
it via limiting constructions. However, the framework's position is
that this compatibility should be understood as \emph{parameterization}
rather than \emph{approximation}.

The sequence $(\mathbb{Z}/n\mathbb{Z})$ does not ``approach'' $S^1$
in the sense that $S^1$ is the ``true'' circle that the discrete
systems are trying to become. Rather, each $\mathbb{Z}/n\mathbb{Z}$
is the circle at resolution $n$ --- a complete, exact geometric object
in its own right. The Gromov--Hausdorff convergence records a
relationship between those objects and the classical space $S^1$ that
arises when the continuum substrate is supplied as an external context.

The framework does not need the limit to define the circle. The limit
is available for those who want to understand the relationship to
classical geometry.
\end{remark}

% ==================================================================
\section{Conclusion}
\label{sec:conclusion}

The central philosophical shift of this framework is that space and
geometric objects are not primitive entities but emergent structures
arising from finite relational processes. We have presented a
process-relational foundation for geometry in which:

\begin{itemize}
  \item relations and compositions are the only primitives,
  \item configurations are finite processes over those relations,
  \item objects are equivalence classes under compositional congruence,
  \item all constructions are strictly finite and discrete.
\end{itemize}

The main theorems establish the core mathematical content.
\Cref{thm:quotient} guarantees that any congruence yields a
well-defined quotient with associative composition.
\Cref{thm:metric} shows that a metric arises from path
optimization over weighted relations, with a discrete corollary
(\Cref{cor:discrete_metric}) when $R$ is finite and weights are
integer-valued.
\Cref{thm:reconstruction} establishes that every finite metric space
arises from a relational system, making the framework universal for
finite metric geometry.
\Cref{thm:equivalence} shows that, after identifying metrically
equivalent relational systems, the metric realization functor is
an equivalence of categories between relational systems and finite
metric spaces.
\Cref{thm:ghbridge} establishes compatibility with classical
continuous geometry via Gromov--Hausdorff convergence, understood
as a bridge result rather than a foundational dependency.
\Cref{thm:pointed_gh} extends this to non-compact spaces including
the $k$-regular tree, using the pointed Gromov--Hausdorff framework.

The examples demonstrate that familiar geometric structures---the
circle, torus, tree, triangulated sphere, and integer grid---have
clean discrete relational representations covering negative, zero,
and positive curvature regimes, all without continuous or infinite
machinery.

\subsection*{Limitations}

We acknowledge several limitations of the present framework. First,
it is not known whether all classical geometric and topological
structures can be expressed as quotients of finite relational systems;
this question is formalized as a future direction below. In particular,
smooth manifolds, Riemannian metrics, and fiber bundles are structures
whose representability within the framework remains open: it is unclear
whether these can be faithfully encoded as finite relational quotients
or whether they require structures that exceed the framework's expressive
capacity. Second, continuous symmetry groups---Lie groups, the
Poincar\'{e} group, $\mathrm{SO}(n)$---do not arise in the framework,
as discussed in Remark~\ref{rem:nosymmetry}; discrete cyclic
approximations are available at any resolution but the continuous groups
themselves are not objects of the theory. Third, the framework handles
static geometric structures; extending it to dynamical settings where
generators and congruences evolve over time is open. These limitations
are consistent with the framework's foundational commitment to finite,
discrete, computable structure: they mark the boundary between what
the framework expresses directly and what requires continuous or
infinite machinery. In each case the limitation is a precise,
identifiable gap rather than a vague incompleteness, and in each case
it points toward a concrete research direction.

Directions for further development:

\begin{enumerate}
  \item \textbf{Classification of congruences.}
        For single-generator systems, the pair $(n, k)$ of
        Proposition~\ref{prop:congclass} is a complete isomorphism
        invariant. For multi-generator systems, characterize which
        congruences produce isomorphic quotients and give a complete
        set of invariants. The natural framework is the correspondence
        between congruences and normal subgroupoids of the free path
        groupoid on $R$, generalizing the subgroup $S \leq
        \mathbb{Z}/n\mathbb{Z}$ of the single-generator case.
  \item \textbf{Reconstruction of discrete classical spaces.} Embed
        known discrete structures (graphs, simplicial complexes, finite
        topological spaces) into the framework as special cases.
  \item \textbf{Higher categorical formalization.} Develop the
        $\omega$-groupoid hierarchy of \Cref{sec:higher} in full detail,
        maintaining finiteness at every level.
  \item \textbf{Observational equivalence.} Determine when two
        relational systems are observationally equivalent---that is,
        produce isomorphic quotients under all congruences---and find
        a complete set of invariants.
  \item \textbf{Applications to discrete physics.} Formalize discrete
        spacetime models (causal sets, spin foams) within the framework
        and characterize their curvature and holonomy relationally.
  \item \textbf{Representability limits.} Identify which classical
        geometric and topological structures --- if any --- cannot be
        expressed as quotients of finite relational systems, and
        characterize the boundary between representable and
        non-representable structures precisely. Understanding this
        boundary is directly relevant to quantum gravity: if certain
        spacetime topologies require structures that exceed what finite
        relational systems can express, that is a substantive constraint
        on any discrete spacetime model, not merely a mathematical
        technicality.
  \item \textbf{Dynamical geometry.} Extend the framework to handle
        relational systems whose generators and congruences evolve ---
        that is, geometry as a process rather than a static structure.
        This would provide a relational account of geometric dynamics,
        relevant to both discrete physics and time-varying computational
        models. In particular, it would allow the framework to address
        questions of curvature change, topological transitions, and the
        evolution of relational structure over time --- currently
        expressible only in continuous settings.
\end{enumerate}

% ==================================================================
\begin{thebibliography}{99}

\bibitem{lawvere}
F.~W.~Lawvere,
``Metric spaces, generalized logic, and closed categories,''
\textit{Rendiconti del Seminario Matematico e Fisico di Milano}
\textbf{43}:135--166, 1973.
\newblock Republished in \textit{Reprints in Theory and Applications of
Categories} \textbf{1} (2002).

\bibitem{buragoetal}
D.~Burago, Y.~Burago, and S.~Ivanov,
\textit{A Course in Metric Geometry},
AMS Graduate Studies in Mathematics \textbf{33},
American Mathematical Society, 2001.
\newblock \doi{10.1090/gsm/033}.

\bibitem{clrs}
T.~H.~Cormen, C.~E.~Leiserson, R.~L.~Rivest, and C.~Stein,
\textit{Introduction to Algorithms},
4th ed., MIT Press, 2022.
\newblock ISBN 978-0-262-04630-5.

\bibitem{knuthbendix}
D.~E.~Knuth and P.~B.~Bendix,
``Simple word problems in universal algebras,''
in J.~Leech (ed.), \textit{Computational Problems in Abstract Algebra},
Pergamon Press, 1970, pp.~263--297.

\bibitem{graphiso}
L.~Babai,
``Graph isomorphism in quasipolynomial time,''
\textit{Proceedings of the 48th ACM Symposium on Theory of Computing
(STOC 2016)}, pp.~684--697, 2016.
\newblock \doi{10.1145/2897518.2897542}.

\bibitem{maclane}
S.~Mac Lane,
\textit{Categories for the Working Mathematician},
2nd ed., Springer, 1998.
\newblock ISBN 978-0-387-98403-2.

\bibitem{awodey}
S.~Awodey,
\textit{Category Theory},
2nd ed., Oxford University Press, 2010.
\newblock \doi{10.1093/acprof:oso/9780198568612.001.0001}.

\bibitem{may}
J.~P.~May,
\textit{A Concise Course in Algebraic Topology},
University of Chicago Press, 1999.
\newblock ISBN 978-0-226-51183-2.

\bibitem{hatcher}
A.~Hatcher,
\textit{Algebraic Topology},
Cambridge University Press, 2002.
\newblock Freely available at \url{https://pi.math.cornell.edu/~hatcher/AT/ATpage.html}.

\bibitem{bookotto}
R.~V.~Book and F.~Otto,
\textit{String-Rewriting Systems},
Springer, 1993.
\newblock \doi{10.1007/978-1-4613-9771-7}.

\bibitem{graphgeom}
R.~Diestel,
\textit{Graph Theory},
5th ed., Springer, 2017.
\newblock \doi{10.1007/978-3-662-53622-3}.

\bibitem{causalset}
L.~Bombelli, J.~Lee, D.~Meyer, and R.~D.~Sorkin,
``Space-time as a causal set,''
\textit{Physical Review Letters} \textbf{59}(5):521--524, 1987.
\newblock \doi{10.1103/PhysRevLett.59.521}.

\end{thebibliography}

\end{document}
